### 2023-09-16 oup-authoring-template.tex
Initial Overleaf Import

%%
%% Copyright 2022 OXFORD UNIVERSITY PRESS
%%
%% This file is part of the 'oup-authoring-template Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'oup-authoring-template Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
%% with bibliographic references
%%

%%%CONTEMPORARY%%%
%\documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
%\documentclass[unnumsec,webpdf,contemporary,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}

%%%MODERN%%%
%\documentclass[unnumsec,webpdf,modern,large]{oup-authoring-template}
\documentclass[unnumsec,webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}

%%%TRADITIONAL%%%
%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}

%\onecolumn % for one column layouts

%\usepackage{showframe}
\usepackage{fontenc}
\graphicspath{{Fig/}}

\usepackage{minted}

% line numbers
%\usepackage[mathlines, switch]{lineno}
%\usepackage[right]{lineno}

\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}

\begin{document}

\journaltitle{Bioinformatics}
\DOI{DOI HERE}
\copyrightyear{2023}
\pubyear{2019}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

\firstpage{1}

%\subtitle{Subject Section}

\title[tstrait]{tstrait: a quantitative trait simulator for ancestral recombination graphs}

\author[1,2,$\ast$]{Daiki Tagami\ORCID{0000-0002-0923-1070}}
\author[2]{Gertjan Bisschop\ORCID{0000-0001-8327-0142}}
\author[2]{Jerome Kelleher\ORCID{0000-0002-7894-5253}}

\authormark{D.Tagami et al.}

\address[1]{\orgdiv{Department of Statistics}, \orgname{University of Oxford}, \orgaddress{\street{24-29 St Giles'}, \postcode{Oxford OX1 3LB}, \country{United Kingdom}}}
\address[2]{\orgdiv{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery}, \orgname{University of Oxford}, \orgaddress{\street{Old Road Campus}, \postcode{Oxford OX3 7LF}, \country{United Kingdom}}}

\corresp[$\ast$]{Corresponding author. \href{email:daiki.tagami@hertford.ox.ac.uk}{daiki.tagami@hertford.ox.ac.uk}}

\received{Date}{0}{Year}
\revised{Date}{0}{Year}
\accepted{Date}{0}{Year}

\abstract{
\textbf{Summary:} 
\texttt{tstrait} is a new open-source Python library that incorporates a wide range of state-of-the-art algorithms for simulating quantitative traits based on an ancestral recombination graph (ARG). ARG is an active field of research in population genetics and statistical genetics due to its promising aspects in genome-wide association study (GWAS), and this package serves as an essential infrastructure for utilizing ARG in GWAS simulations. In this work, we provide an overview of the simulation framework that is incorporated in \texttt{tstrait} and highlight some recent developments in ARG.\\
\textbf{Availability and Implementation:} \texttt{tstrait} is available as an open-source Python package. The full documentation with examples and workflow templates is available on \url{https://tskit.dev/tstrait/docs/}, and the development version is maintained on Github (\url{https://github.com/tskit-dev/tstrait}).\\
\textbf{Contact:} \href{daiki.tagami@hertford.ox.ac.uk}{daiki.tagami@hertford.ox.ac.uk}\\
}
%\abstract{Abstracts must be able to stand alone and so cannot contain %citations to
%the paper's references, equations, etc. An abstract must consist of a single
%paragraph and be concise. Because of online formatting, abstracts must appear
%as plain as possible.}
%\keywords{keyword1, Keyword2, Keyword3, Keyword4}

% \boxedtext{
% \begin{itemize}
% \item Key boxed text here.
% \item Key boxed text here.
% \item Key boxed text here.
% \end{itemize}}

\maketitle

\section{Introduction}

Genome-wide association study (GWAS) tests genetic variants across the entire genome of individuals to identify genetic variants that are statistically associated with a specific trait \citep{uffelmann2021}. GWAS has successfully identified many loci that are associated with various human diseases and traits \citep{ishigaki2022,locke2015,mahajan2022,mathieson2023,yengo2022}, and GWAS results have actively been used in human trials and clinical practices \citep{visscher2017}. In spite of its advantages, population structure and demographic histories can heavily influence GWAS results \citep{uffelmann2021}, so it would be necessary for researchers to implement these factors when conducting GWAS.

Ancestral recombination graphs (ARGs) have many promising aspects in GWAS. Recent studies have shown that genealogy-based association can detect more ultra rare variants than conventional methods \citep{zhang2023}, using an ARG-based approach in quantitative-trait locus mapping can better detect causal locus, especially in the presence of allelic heterogeneity \citep{link2023}, and linkage disequilibirium can be efficiently modeled in GWAS by using a tree-based approach \citep{salehi2023}. These recent advances make ARG to be a suitable data format for conducting GWAS.

ARG is especially useful in population genetic simulations, and it would be possible to simulate genetic data with ancestral information by using ARG. Simulation is used in GWAS to examine how population structure and demographic histories influence polygenic risk score prediction \citep{martin2017,zaidi2020}, because simulation is the only way to obtain genetic data where the underlying demographic information is known. A succinct tree sequence (or tree sequence) is a data structure for ARGs, and it is one of the most popular genomic data structures in population genetic simulations \citep{baumdicker2022,haller2019}, (cite "A general and efficient, ..." article). Since it can efficiently store genetic information with far less data size than other methods, it is an appealing choice for simulating mutation and ancestry information \citep{kelleher2019}.

There are multiple useful software packages to simulate genotypic data with quantitative traits \citep{gaynor2021,haller2023}, and to simulate quantitative traits from observed genetic sequences \citep{fernandes2020,meyer2018}. However, there does not exist any simulators that can efficiently simulate quantitative traits of individuals directly from an ARG. While it would be possible to convert a tree sequence data to a different data format and use the available packages to simulate quantitative traits, data conversion can be computationally costly and it diminishes the computational advantages of using tree sequence data. It would be much more efficient to analyze genetic information by using the tree sequence data structure, and many population genetic statistics can be efficiently computed from the tree structure instead of obtaining the genotype matrix \citep{ralph2020}.

In this paper, we propose a new Python library \texttt{tstrait} to simulate quantitative traits of individuals in an arbitrary ARG. \texttt{tstrait} is built on top of \texttt{tskit} \citep{ralph2020}, which provides an efficient way for analyzing and processing ARGs encoded as tree sequence. \texttt{tstrait} utilizes the efficient data structure in the tree sequence data to simulate complex traits at a fast computational speed. It also provides an easily accessible interface by using \texttt{pandas} dataframe \citep{pandas} as an input and output of the simulation. Thus, \texttt{tstrait} can be easily integrated with other population genetic simulators as well.

In addition to its computational advantages of processing tree sequence data, \texttt{tstrait} offers other advantages: \emph{i)} it can simulate traits even if the sites have multiple mutations, unlike other simulation frameworks that assume infinite-sites model \citep{martin2017}, \emph{ii)} it can simulate traits from any ancestral individuals that existed in the demographic history, unlike other packages that can only simulate traits from the observed genetic information \citep{fernandes2020,meyer2018}, and \emph{iii)} it integrates with Python data science packages, such as \texttt{NumPy} \citep{numpy} and \texttt{pandas} \citep{pandas}. Due to these advantages, we hope that \texttt{tstrait} can provide easily accessible state-of-art quantitative trait simulation for researchers who are using population genetic simulation tools.

\section{Phenotypic Model}

Complex traits are traits that are influenced by a mixture of genetic and environmental factors. The additive model is a common assumption used in GWAS \citep{uffelmann2021}, and \texttt{tstrait} assumes that the traits are obtained from the following phenotypic model,
\begin{align}\label{eq:additive-model}
    y=X\beta+\epsilon,
\end{align}
where $\beta$ is the effect size, $X$ is the matrix that describes the number of causal mutations in the individual, and $\epsilon$ is the environmental noise. The environmental noise is simulated from the following normal distribution,
\begin{align}\label{eq:env}
    \epsilon\sim \mathcal{N}\left(0,\frac{V_G(1-h^2)}{h^2}\right),
\end{align}
where $V_G$ is the additive genetic variance from the simulated effect sizes and $h^2$ is the narrow-sense heritability that is defined by the user.

\section{Package Overview}

\subsection{Effect Size}

The first step of quantitative trait simulation in \texttt{tstrait} is to set the trait model distribution by using \emph{trait\_model} function, as in the following example:
\begin{minted}{python}
model = tstrait.trait_model(
  distribution="normal", mean=0, var=1
)
\end{minted}
The \texttt{tstrait} package supports effect size simulation from five univariate distributions and one multivariate distribution, which are taken from commonly used statistical distributions to simulate effect sizes \citep{gaynor2021,haller2023}. The multivariate normal distribution trait model is used to simulate multiple traits, assuming that they are pleiotropic. The trait model is used as an input of \emph{sim\_trait} function to simulate effect sizes:
\begin{minted}{python}
trait_df = tstrait.sim_trait(
   ts, num_causal=5, model=model
)
\end{minted}
\emph{ts} is the tree sequence input, and the user can set the number of causal sites that is affecting the trait. The output will be a \texttt{pandas} dataframe \citep{pandas} that describes the simulated effect sizes.

\subsection{Genetic Value}

Genetic value is simulated in \texttt{tstrait} by using the effect size dataframe:
\begin{minted}{python}
genetic_result = tstrait.sim_genetic(
   ts, trait_df, alpha=0
)
\end{minted}
\texttt{tstrait} offers flexibility in genetic value simulation, as users can define their own effect size dataframe, in addition to using the simulated output of \emph{sim\_trait} function. This allows users to use the output from other simulators to simulate traits in \texttt{tstrait}.

This is the most computationally intensive part of the simulation algorithm, as it would be necessary for the algorithm to count the number of causal mutations in each individual inside the dataset. The \texttt{sim\_genetic} function uses the tree-traversal algorithm to analyze the tree sequence data \citep{ralph2020}, and uses the Python package \texttt{Numba} \citep{numba} to speed up the code. This enables \texttt{tstrait} to be an efficient quantitative trait simulator of a tree sequence data.

\texttt{tstrait} also supports effect size simulation from the frequency dependent model, where the simulated effect sizes from the causal mutation having allele frequency $p$ will be multiplied by $[2p(1-p)]^\alpha$. The $\alpha$ parameter is a user-defined parameter, which is used to control the degree of frequency dependence on simulated effect sizes. It has been suggested that effect sizes from rarer variants can have increased magnitude from negative selection, and the model above with a negative $\alpha$ value is used to conduct realistic simulations \citep{speed2017}.

\subsection{Phenotype}

Phenotype is simulated from \texttt{tstrait} by using \emph{sim\_env}:
\begin{minted}{python}
phenotype_df = tstrait.sim_env(
   genetic_result.genetic, h2=0.3, random_seed=1
)
\end{minted}
The parameters are used to simulate environmental noise from the distribution (\ref{eq:env}).

After completing the simulation, \texttt{tstrait} returns two dataframes to describe the simulated genetic effect sizes and phenotypes. The resulting dataframes include information regarding the site IDs, individual IDs and allele frequencies, so it would be possible to regenerate the simulated results whenever required.

Alternatively, all the above steps of the simulation can be conducted by using \emph{sim\_phenotype}:
\begin{minted}{python}
sim_result = tstrait.sim_phenotype(
  ts=ts, num_causal=100, model=model, h2=0.3
)
\end{minted}
The user only needs to input the number of causal sites, trait model and the narrow-sense heritability to simulate quantitative traits. By using this function, the user does not need to modify the details of the quantitative trait simulation.

\section{Computational Time}

\texttt{tstrait} offers a very efficient way of simulating quantitative traits based on an ARG. Figure \ref{fig:time} (A) shows how the simulation time scales with the number of individuals on simulated human genomes in \texttt{stdpopsim} \citep{adrion2020}. The only published code to simulate quantitative traits directly from ARGs is in \cite{martin2017}, which we compared with. Including other tools that require the entire genome matrix is not practical. For example, even with 1,000 individuals of 100 Mb human chromosome requires XX Gb of VCF. but it only takes 0.353 seconds to simulate quantitative traits with 1000 causal sites by using \texttt{tstrait}. Thus, the simulation time will be dominated by processing the VCF file. For larger samples, the problem is more

For example, the French Canadian dataset takes XX petabytes in the VCF format, but it is only XX Gb in the \texttt{tskit} compressed tree-sequence format and it only takes a few seconds to load and process by using the tree sequence data structure.

Figure \ref{fig:time} (B) shows simulation time of \texttt{tstrait} with more realistic data. For example, \texttt{tstrait} is used to simulate traits from 2.7 million individuals in the French Canadian dataset, which is the largest known simulated genetic dataset \citep{anderson2023}. It took 60 seconds to simulate a trait with 1000 causal sites, making \texttt{tstrait} to be practical to simulate quantitative traits from this largest ever simulated dataset. The inferred ARG from the 1000 Genomes project \citep{kelleher2019} is also used to measure the computational performance.

\begin{figure}[!t]%
\centering
\includegraphics[width=213pt]{figure.png}
\caption{\textbf{Time taken to simulate quantitative traits.} Each point represents the mean time for 10 independent runs under different random seeds. The times reported are the total CPU time required to simulate quantitative traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM.
The trait model is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and $\alpha=-1$. (A) Computational time with increasing number of individuals. The whole genome dataset from \emph{Homo sapiens} with various number of individuals and chromosome length are simulated by using \texttt{stdpopsim} \citep{adrion2020}. For each replicate of the simulation, \texttt{tstrait} is used to simulate one quantitative trait with 1000 causal sites. (B) Computational time with increasing number of causal sites on two large realistic ARGs. \texttt{tstrait} is used to simulate quantitative traits from the simulated French Canadian dataset \citep{anderson2023} and the inferred tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with varying number of causal sites. The French Canadian dataset is downloaded from \url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000 Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}. Chromosome 9 is selected for both datasets.}\label{fig:time}
\end{figure}

\section{Conclusion}

We believe that in the coming years, simulation in tree sequence encoding will attract more users in need of conducting genetic simulations with complex population structure. Multiple studies highlight the advantages of using the ARG data structure in GWAS \citep{link2023,salehi2023,zhang2023}. ARG data structure is becoming increasingly useful in genetic studies, as it is possible to accurately infer biobank-scale genealogies from sequencing data \citep{zhang2023}, and simulate large sample whole-genome sequences with spatiotemporal metadata \citep{anderson2023}. These advances in ARG presents a real opportunity for people to further understand the effects of complex population structure on quantitative traits. The \texttt{tstrait} package is an essential piece of infrastructure that lets people explore these possibilities.

\section{Competing interests}
No competing interest is declared.

\section{Acknowledgments}

We greatly acknowledge helpful discussions about quantitative trait simulation with Gregor Gorjanc. We thank Ben Haller and people in the tskit community for helpful discussions and feedback. We thank Ben Jeffery for his help with the Python package development.

\section{Funding}

This work was supported by ...

%USE THE BELOW OPTIONS IN CASE YOU NEED AUTHOR YEAR FORMAT.
\bibliographystyle{abbrvnat}
\bibliography{reference}

\end{document}

### 2023-09-16 computational-time/genotype_matrix.py
Code + data

@@ -1,0 +1,51 @@
+# Mertin et al. (2017) simulation
+import numpy as np
+from dataclasses import dataclass
+
+@dataclass
+class Result:
+    phenotype: np.ndarray
+    site_id: np.ndarray
+    causal_allele: np.ndarray
+    effect_size: np.ndarray
+    allele_frequency: np.ndarray    
+
+
+def matrix(ts, h2, ncausal, random_seed):
+    
+    rng = np.random.default_rng(random_seed)
+    
+    causal_site_index = np.linspace(0, ts.num_sites-1, ncausal, dtype=int)
+    
+    allele_freq = np.zeros(ncausal, dtype=float)
+    effect_size = np.zeros(ncausal, dtype=float)
+    causal_allele = np.zeros(ncausal, dtype=object)
+    
+    nhaps = ts.num_samples
+
+    prs_haps = np.zeros(nhaps) #score for each haplotype
+    i = 0
+    for variant in ts.variants():
+        if variant.index in causal_site_index:
+            sim_effect_size = np.random.normal(loc=0,scale=h2/ncausal)
+            effect_size[i] = sim_effect_size
+            causal_allele_index = rng.integers(low=1, high=len(variant.alleles))
+            allele = variant.alleles[causal_allele_index]
+            causal_allele[i] = allele
+            genotype = np.array([1 if i == causal_allele_index else 0 for i in variant.genotypes])
+            prs_haps += genotype * sim_effect_size
+            allele_freq[i] = np.sum(genotype) / len(genotype)
+            i += 1
+
+    env_effect = rng.normal(loc=0,scale=1-h2, size=nhaps)
+    prs_norm = (prs_haps - np.mean(prs_haps)) / np.std(prs_haps)
+    env_norm = (env_effect - np.mean(env_effect)) / np.std(env_effect)
+    total_liability = np.sqrt(h2) * prs_norm + np.sqrt(1 - h2) * env_norm
+    
+    simulation_result = Result(phenotype = total_liability,
+                                        site_id = causal_site_index,
+                                       causal_allele = causal_allele,
+                                       effect_size = effect_size,
+                                       allele_frequency = allele_freq)
+    
+    return simulation_result

### 2023-09-16 oup-authoring-template.tex

@@
 The only published code to simulate quantitative traits directly from ARGs is in \cite{martin2017}, which we compared with.
-Including other tools that require the entire genome matrix is not practical.
-For example, even with 1,000 individuals of 100 Mb human chromosome requires XX Gb of VCF.
-but it only takes 0.353 seconds to simulate quantitative traits with 1000 causal sites by using \texttt{tstrait}.
-Thus, the simulation time will be dominated by processing the VCF file.
-For larger samples, the problem is more
-
-For example, the French Canadian dataset takes XX petabytes in the VCF format, but it is only XX Gb in the \texttt{tskit} compressed tree-sequence format and it only takes a few seconds to load and process by using the tree sequence data structure.
+Including other tools that require the entire genome matrix is not practical, as the simulation time will be dominated by processing the VCF file and it would not be practically to analyze a large sample VCF file on a laptop PC.
+For example, the Chromosome 9 French Canadian dataset takes approximately 280 terabytes in the VCF format, but it is only 1.357 Gb in the \texttt{tskit} compressed tree-sequence format and it only takes a few minutes to load and process by using the tree sequence data structure.
 
@@
 
+D.T. is supported by the Oxford Kobe Scholarship from the University of Oxford and the Euretta J. Kellett Fellowship from Columbia University.
+
 This work was supported by ...

### 2023-10-12 cover_letter/cover_letter.tex
Create cover_letter.tex

\documentclass{letter}

\signature{Daiki Tagami}

\address{Big Data Institute\\University of Oxford\\UK}
\begin{document}

\begin{letter}{Bioinformatics}

\opening{Dear Editors,}

I am writing on behalf of my coauthors to submit our manuscript entitled
\emph{tstrait: a quantitative trait simulator for ancestral recombination graphs}
for consideration for publication in Bioinformatics
as an ``Application Note'' research article.

\texttt{tstrait} is a newly developed Python package to simulate quantitative traits from an ancestral recombination graph (ARG). ARG is widely used in population genetics and statistical genetics due to its advantages in genome-wide association study (GWAS), but there does not exist any simulators that can efficiently simulate quantitative traits directly from an ARG. \texttt{tstrait} can simulate quantitative traits at a substantially faster computational speed than traditional techniques, making it to be an essential piece of infrastructure in ARG and GWAS.

We believe that this manuscript provides an overview of current works in ARG and GWAS, and presents the importance of using \texttt{tstrait} in genomic studies. It will be of substantial interest to the readers of Bioinformatics, and we hope that you agree.

\closing{Sincerely,}

\end{letter}
\end{document}

### 2023-11-06 paper.tex
Rename
Line breaks

--- 2023-09-16 oup-authoring-template.tex
@@
 \usepackage{minted}
+
+\usepackage{graphicx}
 
@@
 \textbf{Summary:}
-\texttt{tstrait} is a new open-source Python library that incorporates a wide range of state-of-the-art algorithms for simulating quantitative traits based on an ancestral recombination graph (ARG).
-ARG is an active field of research in population genetics and statistical genetics due to its promising aspects in genome-wide association study (GWAS), and this package serves as an essential infrastructure for utilizing ARG in GWAS simulations.
-In this work, we provide an overview of the simulation framework that is incorporated in \texttt{tstrait} and highlight some recent developments in ARG.\\
-\textbf{Availability and Implementation:} \texttt{tstrait} is available as an open-source Python package.
-The full documentation with examples and workflow templates is available on \url{https://tskit.dev/tstrait/docs/}, and the development version is maintained on Github (\url{https://github.com/tskit-dev/tstrait}).\\
-\textbf{Contact:} \href{daiki.tagami@hertford.ox.ac.uk}{daiki.tagami@hertford.ox.ac.uk}\\
+\texttt{tstrait} is a new open-source Python library that incorporates a wide
+range of state-of-the-art algorithms for simulating quantitative traits based
+on an ancestral recombination graph (ARG). ARG is an active field of research
+in population genetics and statistical genetics due to its promising aspects in
+genome-wide association study (GWAS), and this package serves as an essential
+infrastructure for utilizing ARG in GWAS simulations. In this work, we provide
+an overview of the simulation framework that is incorporated in
+\texttt{tstrait} and highlight some recent developments in ARG.\\
+\textbf{Availability and Implementation:} \texttt{tstrait} is available as an
+open-source Python package. The full documentation with examples and workflow
+templates is available on \url{https://tskit.dev/tstrait/docs/}, and the
+development version is maintained on Github
+(\url{https://github.com/tskit-dev/tstrait}).\\ \textbf{Contact:}
+\href{daiki.tagami@hertford.ox.ac.uk}{daiki.tagami@hertford.ox.ac.uk}\\
 }
@@
 
-Genome-wide association study (GWAS) tests genetic variants across the entire genome of individuals to identify genetic variants that are statistically associated with a specific trait \citep{uffelmann2021}.
-GWAS has successfully identified many loci that are associated with various human diseases and traits \citep{ishigaki2022,locke2015,mahajan2022,mathieson2023,yengo2022}, and GWAS results have actively been used in human trials and clinical practices \citep{visscher2017}.
-In spite of its advantages, population structure and demographic histories can heavily influence GWAS results \citep{uffelmann2021}, so it would be necessary for researchers to implement these factors when conducting GWAS.
+Genome-wide association study (GWAS) tests genetic variants across the entire
+genome of individuals to identify genetic variants that are statistically
+associated with a specific trait \citep{uffelmann2021}. GWAS has successfully
+identified many loci that are associated with various human diseases and traits
+\citep{ishigaki2022,locke2015,mahajan2022,mathieson2023,yengo2022}, and GWAS
+results have actively been used in human trials and clinical practices
+\citep{visscher2017}. In spite of its advantages, population structure and
+demographic histories can heavily influence GWAS results \citep{uffelmann2021},
+so it would be necessary for researchers to implement these factors when
+conducting GWAS.
 
 Ancestral recombination graphs (ARGs) have many promising aspects in GWAS.
-Recent studies have shown that genealogy-based association can detect more ultra rare variants than conventional methods \citep{zhang2023}, using an ARG-based approach in quantitative-trait locus mapping can better detect causal locus, especially in the presence of allelic heterogeneity \citep{link2023}, and linkage disequilibirium can be efficiently modeled in GWAS by using a tree-based approach \citep{salehi2023}.
-These recent advances make ARG to be a suitable data format for conducting GWAS.
+Recent studies have shown that genealogy-based association can detect more
+ultra rare variants than conventional methods \citep{zhang2023}, using an
+ARG-based approach in quantitative-trait locus mapping can better detect causal
+locus, especially in the presence of allelic heterogeneity \citep{link2023},
+and linkage disequilibirium can be efficiently modeled in GWAS by using a
+tree-based approach \citep{salehi2023}. These recent advances make ARG to be a
+suitable data format for conducting GWAS.
 
-ARG is especially useful in population genetic simulations, and it would be possible to simulate genetic data with ancestral information by using ARG.
-Simulation is used in GWAS to examine how population structure and demographic histories influence polygenic risk score prediction \citep{martin2017,zaidi2020}, because simulation is the only way to obtain genetic data where the underlying demographic information is known.
-A succinct tree sequence (or tree sequence) is a data structure for ARGs, and it is one of the most popular genomic data structures in population genetic simulations \citep{baumdicker2022,haller2019}, (cite "A general and efficient, ..." article).
-Since it can efficiently store genetic information with far less data size than other methods, it is an appealing choice for simulating mutation and ancestry information \citep{kelleher2019}.
+ARG is especially useful in population genetic simulations, and it would be
+possible to simulate genetic data with ancestral information by using ARG.
+Simulation is used in GWAS to examine how population structure and demographic
+histories influence polygenic risk score prediction
+\citep{martin2017,zaidi2020}, because simulation is the only way to obtain
+genetic data where the underlying demographic information is known. A succinct
+tree sequence (or tree sequence) is a data structure for ARGs, and it is one of
+the most popular genomic data structures in population genetic simulations
+\citep{baumdicker2022,haller2019}, (cite "A general and efficient, ..."
+article). Since it can efficiently store genetic information with far less data
+size than other methods, it is an appealing choice for simulating mutation and
+ancestry information \citep{kelleher2019}.
 
-There are multiple useful software packages to simulate genotypic data with quantitative traits \citep{gaynor2021,haller2023}, and to simulate quantitative traits from observed genetic sequences \citep{fernandes2020,meyer2018}.
-However, there does not exist any simulators that can efficiently simulate quantitative traits of individuals directly from an ARG.
-While it would be possible to convert a tree sequence data to a different data format and use the available packages to simulate quantitative traits, data conversion can be computationally costly and it diminishes the computational advantages of using tree sequence data.
-It would be much more efficient to analyze genetic information by using the tree sequence data structure, and many population genetic statistics can be efficiently computed from the tree structure instead of obtaining the genotype matrix \citep{ralph2020}.
+There are multiple useful software packages to simulate genotypic data with
+quantitative traits \citep{gaynor2021,haller2023}, and to simulate quantitative
+traits from observed genetic sequences \citep{fernandes2020,meyer2018}.
+However, there does not exist any simulators that can efficiently simulate
+quantitative traits of individuals directly from an ARG. While it would be
+possible to convert a tree sequence data to a different data format and use the
+available packages to simulate quantitative traits, data conversion can be
+computationally costly and it diminishes the computational advantages of using
+tree sequence data. It would be much more efficient to analyze genetic
+information by using the tree sequence data structure, and many population
+genetic statistics can be efficiently computed from the tree structure instead
+of obtaining the genotype matrix \citep{ralph2020}.
 
-In this paper, we propose a new Python library \texttt{tstrait} to simulate quantitative traits of individuals in an arbitrary ARG.
-\texttt{tstrait} is built on top of \texttt{tskit} \citep{ralph2020}, which provides an efficient way for analyzing and processing ARGs encoded as tree sequence.
-\texttt{tstrait} utilizes the efficient data structure in the tree sequence data to simulate complex traits at a fast computational speed.
-It also provides an easily accessible interface by using \texttt{pandas} dataframe \citep{pandas} as an input and output of the simulation.
-Thus, \texttt{tstrait} can be easily integrated with other population genetic simulators as well.
+In this paper, we propose a new Python library \texttt{tstrait} to simulate
+quantitative traits of individuals in an arbitrary ARG. \texttt{tstrait} is
+built on top of \texttt{tskit} \citep{ralph2020}, which provides an efficient
+way for analyzing and processing ARGs encoded as tree sequence.
+\texttt{tstrait} utilizes the efficient data structure in the tree sequence
+data to simulate complex traits at a fast computational speed. It also provides
+an easily accessible interface by using \texttt{pandas} dataframe
+\citep{pandas} as an input and output of the simulation. Thus, \texttt{tstrait}
+can be easily integrated with other population genetic simulators as well.
 
-In addition to its computational advantages of processing tree sequence data, \texttt{tstrait} offers other advantages: \emph{i)} it can simulate traits even if the sites have multiple mutations, unlike other simulation frameworks that assume infinite-sites model \citep{martin2017}, \emph{ii)} it can simulate traits from any ancestral individuals that existed in the demographic history, unlike other packages that can only simulate traits from the observed genetic information \citep{fernandes2020,meyer2018}, and \emph{iii)} it integrates with Python data science packages, such as \texttt{NumPy} \citep{numpy} and \texttt{pandas} \citep{pandas}.
-Due to these advantages, we hope that \texttt{tstrait} can provide easily accessible state-of-art quantitative trait simulation for researchers who are using population genetic simulation tools.
+In addition to its computational advantages of processing tree sequence data,
+\texttt{tstrait} offers other advantages: \emph{i)} it can simulate traits even
+if the sites have multiple mutations, unlike other simulation frameworks that
+assume infinite-sites model \citep{martin2017}, \emph{ii)} it can simulate
+traits from any ancestral individuals that existed in the demographic history,
+unlike other packages that can only simulate traits from the observed genetic
+information \citep{fernandes2020,meyer2018}, and \emph{iii)} it integrates with
+Python data science packages, such as \texttt{NumPy} \citep{numpy} and
+\texttt{pandas} \citep{pandas}. Due to these advantages, we hope that
+\texttt{tstrait} can provide easily accessible state-of-art quantitative trait
+simulation for researchers who are using population genetic simulation tools.
 
@@
 
-Complex traits are traits that are influenced by a mixture of genetic and environmental factors.
-The additive model is a common assumption used in GWAS \citep{uffelmann2021}, and \texttt{tstrait} assumes that the traits are obtained from the following phenotypic model,
+Complex traits are traits that are influenced by a mixture of genetic and
+environmental factors. The additive model is a common assumption used in GWAS
+\citep{uffelmann2021}, and \texttt{tstrait} assumes that the traits are
+obtained from the following phenotypic model,
 \begin{align}\label{eq:additive-model}
@@
 \end{align}
-where $\beta$ is the effect size, $X$ is the matrix that describes the number of causal mutations in the individual, and $\epsilon$ is the environmental noise.
-The environmental noise is simulated from the following normal distribution,
+where $\beta$ is the effect size, $X$ is the matrix that describes the number
+of causal mutations in the individual, and $\epsilon$ is the environmental
+noise. The environmental noise is simulated from the following normal
+distribution,
 \begin{align}\label{eq:env}
@@
 \end{align}
-where $V_G$ is the additive genetic variance from the simulated effect sizes and $h^2$ is the narrow-sense heritability that is defined by the user.
+where $V_G$ is the additive genetic variance from the simulated effect sizes
+and $h^2$ is the narrow-sense heritability that is defined by the user.
 
@@
 
-The first step of quantitative trait simulation in \texttt{tstrait} is to set the trait model distribution by using \emph{trait\_model} function, as in the following example:
+The first step of quantitative trait simulation in \texttt{tstrait} is to set
+the trait model distribution by using \emph{trait\_model} function, as in the
+following example:
 \begin{minted}{python}
@@
 \end{minted}
-The \texttt{tstrait} package supports effect size simulation from five univariate distributions and one multivariate distribution, which are taken from commonly used statistical distributions to simulate effect sizes \citep{gaynor2021,haller2023}.
-The multivariate normal distribution trait model is used to simulate multiple traits, assuming that they are pleiotropic.
-The trait model is used as an input of \emph{sim\_trait} function to simulate effect sizes:
+The \texttt{tstrait} package supports effect size simulation from five
+univariate distributions and one multivariate distribution, which are taken
+from commonly used statistical distributions to simulate effect sizes
+\citep{gaynor2021,haller2023}. The multivariate normal distribution trait model
+is used to simulate multiple traits, assuming that they are pleiotropic. The
+trait model is used as an input of \emph{sim\_trait} function to simulate
+effect sizes:
 \begin{minted}{python}
@@
 \end{minted}
-\emph{ts} is the tree sequence input, and the user can set the number of causal sites that is affecting the trait.
-The output will be a \texttt{pandas} dataframe \citep{pandas} that describes the simulated effect sizes.
+\emph{ts} is the tree sequence input, and the user can set the number of causal
+sites that is affecting the trait. The output will be a \texttt{pandas}
+dataframe \citep{pandas} that describes the simulated effect sizes.
 
@@
 \end{minted}
-\texttt{tstrait} offers flexibility in genetic value simulation, as users can define their own effect size dataframe, in addition to using the simulated output of \emph{sim\_trait} function.
-This allows users to use the output from other simulators to simulate traits in \texttt{tstrait}.
+\texttt{tstrait} offers flexibility in genetic value simulation, as users can
+define their own effect size dataframe, in addition to using the simulated
+output of \emph{sim\_trait} function. This allows users to use the output from
+other simulators to simulate traits in \texttt{tstrait}.
 
-This is the most computationally intensive part of the simulation algorithm, as it would be necessary for the algorithm to count the number of causal mutations in each individual inside the dataset.
-The \texttt{sim\_genetic} function uses the tree-traversal algorithm to analyze the tree sequence data \citep{ralph2020}, and uses the Python package \texttt{Numba} \citep{numba} to speed up the code.
-This enables \texttt{tstrait} to be an efficient quantitative trait simulator of a tree sequence data.
+This is the most computationally intensive part of the simulation algorithm, as
+it would be necessary for the algorithm to count the number of causal mutations
+in each individual inside the dataset. The \texttt{sim\_genetic} function uses
+the tree-traversal algorithm to analyze the tree sequence data
+\citep{ralph2020}, and uses the Python package \texttt{Numba} \citep{numba} to
+speed up the code. This enables \texttt{tstrait} to be an efficient
+quantitative trait simulator of a tree sequence data.
 
-\texttt{tstrait} also supports effect size simulation from the frequency dependent model, where the simulated effect sizes from the causal mutation having allele frequency $p$ will be multiplied by $[2p(1-p)]^\alpha$.
-The $\alpha$ parameter is a user-defined parameter, which is used to control the degree of frequency dependence on simulated effect sizes.
-It has been suggested that effect sizes from rarer variants can have increased magnitude from negative selection, and the model above with a negative $\alpha$ value is used to conduct realistic simulations \citep{speed2017}.
+\texttt{tstrait} also supports effect size simulation from the frequency
+dependent model, where the simulated effect sizes from the causal mutation
+having allele frequency $p$ will be multiplied by $[2p(1-p)]^\alpha$. The
+$\alpha$ parameter is a user-defined parameter, which is used to control the
+degree of frequency dependence on simulated effect sizes. It has been suggested
+that effect sizes from rarer variants can have increased magnitude from
+negative selection, and the model above with a negative $\alpha$ value is used
+to conduct realistic simulations \citep{speed2017}.
 
@@
 
-After completing the simulation, \texttt{tstrait} returns two dataframes to describe the simulated genetic effect sizes and phenotypes.
-The resulting dataframes include information regarding the site IDs, individual IDs and allele frequencies, so it would be possible to regenerate the simulated results whenever required.
+After completing the simulation, \texttt{tstrait} returns two dataframes to
+describe the simulated genetic effect sizes and phenotypes. The resulting
+dataframes include information regarding the site IDs, individual IDs and
+allele frequencies, so it would be possible to regenerate the simulated results
+whenever required.
 
-Alternatively, all the above steps of the simulation can be conducted by using \emph{sim\_phenotype}:
+Alternatively, all the above steps of the simulation can be conducted by using
+\emph{sim\_phenotype}:
 \begin{minted}{python}
@@
 \end{minted}
-The user only needs to input the number of causal sites, trait model and the narrow-sense heritability to simulate quantitative traits.
-By using this function, the user does not need to modify the details of the quantitative trait simulation.
+The user only needs to input the number of causal sites, trait model and the
+narrow-sense heritability to simulate quantitative traits. By using this
+function, the user does not need to modify the details of the quantitative
+trait simulation.
 
@@
 
-\texttt{tstrait} offers a very efficient way of simulating quantitative traits based on an ARG.
-Figure \ref{fig:time} (A) shows how the simulation time scales with the number of individuals on simulated human genomes in \texttt{stdpopsim} \citep{adrion2020}.
-The only published code to simulate quantitative traits directly from ARGs is in \cite{martin2017}, which we compared with.
-Including other tools that require the entire genome matrix is not practical, as the simulation time will be dominated by processing the VCF file and it would not be practically to analyze a large sample VCF file on a laptop PC.
-For example, the Chromosome 9 French Canadian dataset takes approximately 280 terabytes in the VCF format, but it is only 1.357 Gb in the \texttt{tskit} compressed tree-sequence format and it only takes a few minutes to load and process by using the tree sequence data structure.
+\texttt{tstrait} offers a very efficient way of simulating quantitative traits
+based on an ARG. Figure \ref{fig:time} (A) shows how the simulation time scales
+with the number of individuals on simulated human genomes in \texttt{stdpopsim}
+\citep{adrion2020}. The only published code to simulate quantitative traits
+directly from ARGs is in \cite{martin2017}, which we compared with. Including
+other tools that require the entire genome matrix is not practical, as the
+simulation time will be dominated by processing the VCF file and it would not
+be practically to analyze a large sample VCF file on a laptop PC. For example,
+the Chromosome 9 French Canadian dataset takes approximately 280 terabytes in
+the VCF format, but it is only 1.357 Gb in the \texttt{tskit} compressed
+tree-sequence format and it only takes a few minutes to load and process by
+using the tree sequence data structure.
 
-Figure \ref{fig:time} (B) shows simulation time of \texttt{tstrait} with more realistic data.
-For example, \texttt{tstrait} is used to simulate traits from 2.7 million individuals in the French Canadian dataset, which is the largest known simulated genetic dataset \citep{anderson2023}.
-It took 60 seconds to simulate a trait with 1000 causal sites, making \texttt{tstrait} to be practical to simulate quantitative traits from this largest ever simulated dataset.
-The inferred ARG from the 1000 Genomes project \citep{kelleher2019} is also used to measure the computational performance.
+Figure \ref{fig:time} (B) shows simulation time of \texttt{tstrait} with more
+realistic data. For example, \texttt{tstrait} is used to simulate traits from
+2.7 million individuals in the French Canadian dataset, which is the largest
+known simulated genetic dataset \citep{anderson2023}. It took 60 seconds to
+simulate a trait with 1000 causal sites, making \texttt{tstrait} to be
+practical to simulate quantitative traits from this largest ever simulated
+dataset. The inferred ARG from the 1000 Genomes project \citep{kelleher2019} is
+also used to measure the computational performance.
 
@@
 \centering
-\includegraphics[width=213pt]{figure.png}
-\caption{\textbf{Time taken to simulate quantitative traits.} Each point represents the mean time for 10 independent runs under different random seeds.
-The times reported are the total CPU time required to simulate quantitative traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM.
-The trait model is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and $\alpha=-1$.
-(A) Computational time with increasing number of individuals.
-The whole genome dataset from \emph{Homo sapiens} with various number of individuals and chromosome length are simulated by using \texttt{stdpopsim} \citep{adrion2020}.
-For each replicate of the simulation, \texttt{tstrait} is used to simulate one quantitative trait with 1000 causal sites.
-(B) Computational time with increasing number of causal sites on two large realistic ARGs.
-\texttt{tstrait} is used to simulate quantitative traits from the simulated French Canadian dataset \citep{anderson2023} and the inferred tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with varying number of causal sites.
-The French Canadian dataset is downloaded from \url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000 Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}.
+\includegraphics[width=213pt]{figure.pdf}
+\caption{\textbf{Time taken to simulate quantitative traits.} Each point
+represents the mean time for 10 independent runs under different random seeds.
+The times reported are the total CPU time required to simulate quantitative
+traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
+is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
+$\alpha=-1$. (A) Computational time with increasing number of individuals. The
+whole genome dataset from \emph{Homo sapiens} with various number of
+individuals and chromosome length are simulated by using \texttt{stdpopsim}
+\citep{adrion2020}. For each replicate of the simulation, \texttt{tstrait} is
+used to simulate one quantitative trait with 1000 causal sites. (B)
+Computational time with increasing number of causal sites on two large
+realistic ARGs. \texttt{tstrait} is used to simulate quantitative traits from
+the simulated French Canadian dataset \citep{anderson2023} and the inferred
+tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with
+varying number of causal sites. The French Canadian dataset is downloaded from
+\url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000
+Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}.
 Chromosome 9 is selected for both datasets.}\label{fig:time}
@@
 
-We believe that in the coming years, simulation in tree sequence encoding will attract more users in need of conducting genetic simulations with complex population structure.
-Multiple studies highlight the advantages of using the ARG data structure in GWAS \citep{link2023,salehi2023,zhang2023}.
-ARG data structure is becoming increasingly useful in genetic studies, as it is possible to accurately infer biobank-scale genealogies from sequencing data \citep{zhang2023}, and simulate large sample whole-genome sequences with spatiotemporal metadata \citep{anderson2023}.
-These advances in ARG presents a real opportunity for people to further understand the effects of complex population structure on quantitative traits.
-The \texttt{tstrait} package is an essential piece of infrastructure that lets people explore these possibilities.
+We believe that in the coming years, simulation in tree sequence encoding will
+attract more users in need of conducting genetic simulations with complex
+population structure. Multiple studies highlight the advantages of using the
+ARG data structure in GWAS \citep{link2023,salehi2023,zhang2023}. ARG data
+structure is becoming increasingly useful in genetic studies, as it is possible
+to accurately infer biobank-scale genealogies from sequencing data
+\citep{zhang2023}, and simulate large sample whole-genome sequences with
+spatiotemporal metadata \citep{anderson2023}. These advances in ARG presents a
+real opportunity for people to further understand the effects of complex
+population structure on quantitative traits. The \texttt{tstrait} package is an
+essential piece of infrastructure that lets people explore these possibilities.
 
@@
 
-We greatly acknowledge helpful discussions about quantitative trait simulation with Gregor Gorjanc.
-We thank Ben Haller and people in the tskit community for helpful discussions and feedback.
-We thank Ben Jeffery for his help with the Python package development.
+We greatly acknowledge helpful discussions about quantitative trait simulation
+with Gregor Gorjanc. We thank Ben Haller and people in the tskit community for
+helpful discussions and feedback. We thank Ben Jeffery for his help with the
+Python package development.
 
@@
 
-D.T. is supported by the Oxford Kobe Scholarship from the University of Oxford and the Euretta J. Kellett Fellowship from Columbia University.
+D.T. is supported by the Oxford Kobe Scholarship from the University of Oxford
+and the Euretta J. Kellett Fellowship from Columbia University.
 
@@
 \bibliographystyle{abbrvnat}
-\bibliography{reference}
+\bibliography{paper}
 

### 2023-11-07 paper.tex
Rejig and remove template files
Switch to lstlisting following template

@@
 \usepackage{fontenc}
-\graphicspath{{Fig/}}
-
-\usepackage{minted}
 
@@
 % line numbers
-%\usepackage[mathlines, switch]{lineno}
-%\usepackage[right]{lineno}
-
-\theoremstyle{thmstyleone}%
-\newtheorem{theorem}{Theorem}%  meant for continuous numbers
-%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
-%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
-\newtheorem{proposition}[theorem]{Proposition}%
-%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
-\theoremstyle{thmstyletwo}%
-\newtheorem{example}{Example}%
-\newtheorem{remark}{Remark}%
-\theoremstyle{thmstylethree}%
-\newtheorem{definition}{Definition}
+% \usepackage[mathlines, switch]{lineno}
+% \usepackage[right]{lineno}
 
@@
 
-\journaltitle{Bioinformatics}
-\DOI{DOI HERE}
+% \journaltitle{Bioinformatics}
+\journaltitle{Preprint}
+% \DOI{DOI HERE}
 \copyrightyear{2023}
-\pubyear{2019}
-\access{Advance Access Publication Date: Day Month Year}
+% \pubyear{2019}
+% \access{Advance Access Publication Date: Day Month Year}
 \appnotes{Paper}
@@
 
-\address[1]{\orgdiv{Department of Statistics}, \orgname{University of Oxford}, \orgaddress{\street{24-29 St Giles'}, \postcode{Oxford OX1 3LB}, \country{United Kingdom}}}
-\address[2]{\orgdiv{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery}, \orgname{University of Oxford}, \orgaddress{\street{Old Road Campus}, \postcode{Oxford OX3 7LF}, \country{United Kingdom}}}
+\address[1]{\orgdiv{Department of Statistics}, \orgname{University of Oxford},
+\orgaddress{\street{24-29 St Giles'}, \postcode{Oxford OX1 3LB},
+\country{United Kingdom}}}
+\address[2]{\orgdiv{Big Data Institute,
+Li Ka Shing Centre for Health Information and Discovery},
+\orgname{University of Oxford}, \orgaddress{\street{Old Road Campus},
+\postcode{Oxford OX3 7LF}, \country{United Kingdom}}}
 
-\corresp[$\ast$]{Corresponding author. \href{email:daiki.tagami@hertford.ox.ac.uk}{daiki.tagami@hertford.ox.ac.uk}}
+\corresp[$\ast$]{Corresponding author.
+\href{email:daiki.tagami@hertford.ox.ac.uk}{daiki.tagami@hertford.ox.ac.uk}}
 
@@
 }
+
 %\abstract{Abstracts must be able to stand alone and so cannot contain %citations to
@@
 %\keywords{keyword1, Keyword2, Keyword3, Keyword4}
-
-% \boxedtext{
-% \begin{itemize}
-% \item Key boxed text here.
-% \item Key boxed text here.
-% \item Key boxed text here.
-% \end{itemize}}
 
@@
 following example:
-\begin{minted}{python}
+\begin{lstlisting}[language=Python]
 model = tstrait.trait_model(
-  distribution="normal", mean=0, var=1
+  distribution='normal', mean=0, var=1
 )
-\end{minted}
+\end{lstlisting}
 The \texttt{tstrait} package supports effect size simulation from five
@@
 effect sizes:
-\begin{minted}{python}
+\begin{lstlisting}[language=Python]
 trait_df = tstrait.sim_trait(
@@
 )
-\end{minted}
+\end{lstlisting}
 \emph{ts} is the tree sequence input, and the user can set the number of causal
@@
 Genetic value is simulated in \texttt{tstrait} by using the effect size dataframe:
-\begin{minted}{python}
+\begin{lstlisting}[language=Python]
 genetic_result = tstrait.sim_genetic(
@@
 )
-\end{minted}
+\end{lstlisting}
 \texttt{tstrait} offers flexibility in genetic value simulation, as users can
@@
 Phenotype is simulated from \texttt{tstrait} by using \emph{sim\_env}:
-\begin{minted}{python}
+\begin{lstlisting}[language=Python]
 phenotype_df = tstrait.sim_env(
@@
 )
-\end{minted}
+\end{lstlisting}
 The parameters are used to simulate environmental noise from the distribution (\ref{eq:env}).
@@
 \emph{sim\_phenotype}:
-\begin{minted}{python}
+\begin{lstlisting}[language=Python]
 sim_result = tstrait.sim_phenotype(
-  ts=ts, num_causal=100, model=model, h2=0.3
+  ts, num_causal=100, model=model, h2=0.3
 )
-\end{minted}
+\end{lstlisting}
 The user only needs to input the number of causal sites, trait model and the
@@
 \centering
-\includegraphics[width=213pt]{figure.pdf}
+\includegraphics[width=213pt]{figures/time-scaling.pdf}
 \caption{\textbf{Time taken to simulate quantitative traits.} Each point

### 2023-11-07 paper.tex
Pass through abstract

@@
 \textbf{Summary:}
-\texttt{tstrait} is a new open-source Python library that incorporates a wide
-range of state-of-the-art algorithms for simulating quantitative traits based
-on an ancestral recombination graph (ARG). ARG is an active field of research
-in population genetics and statistical genetics due to its promising aspects in
-genome-wide association study (GWAS), and this package serves as an essential
-infrastructure for utilizing ARG in GWAS simulations. In this work, we provide
-an overview of the simulation framework that is incorporated in
-\texttt{tstrait} and highlight some recent developments in ARG.\\
-\textbf{Availability and Implementation:} \texttt{tstrait} is available as an
-open-source Python package. The full documentation with examples and workflow
+Ancestral recombination graphs (ARGs) encode the entire ensemble of correlated
+genealogical trees arising from recombination in a compact and efficient
+structure, and are of fundamental importance in population and statistical
+genetics. Recent breakthroughs have made it possible to simulate and
+infer ARGs at the population scale, and there is now intense interest
+in applying ARG-based methods across a broad range of applications.
+Genome-wide association studies (GWAS) are one of the most imporant
+application areas, with ARG-based methods already providing substantial
+improvements over the state-of-the-art. The ability to efficiently
+simulate quantitative traits is critical to developing
+and evaluating new GWAS methods, but existing simulators do not scale
+to modern datasets and cannot be directly applied to inferred or simulated ARGs.
+We present the \texttt{tstrait} package,
+an open-source Python library to simulate
+quantitative traits on ARGs, and show how this user-friendly software
+can be easily applied to vast, population-scale datasets.
+\\\textbf{Availability and Implementation:} \texttt{tstrait} is available
+for download on the Python Package Index.
+Full documentation with examples and workflow
 templates is available on \url{https://tskit.dev/tstrait/docs/}, and the

### 2023-11-07 paper.tex
Fixup bib

@@
 \textbf{Summary:}
-Ancestral recombination graphs (ARGs) encode the entire ensemble of correlated
+Ancestral recombination graphs (ARGs) encode the ensemble of correlated
 genealogical trees arising from recombination in a compact and efficient
@@
 and evaluating new GWAS methods, but existing simulators do not scale
-to modern datasets and cannot be directly applied to inferred or simulated ARGs.
+to modern datasets and cannot be directly applied to ARGs.
 We present the \texttt{tstrait} package,
@@
 %as plain as possible.}
-%\keywords{keyword1, Keyword2, Keyword3, Keyword4}
+\keywords{Ancestral Recombination Graph, ARG, GWAS, quantitative traits}
 
@@
 
-Genome-wide association study (GWAS) tests genetic variants across the entire
-genome of individuals to identify genetic variants that are statistically
-associated with a specific trait \citep{uffelmann2021}. GWAS has successfully
-identified many loci that are associated with various human diseases and traits
-\citep{ishigaki2022,locke2015,mahajan2022,mathieson2023,yengo2022}, and GWAS
-results have actively been used in human trials and clinical practices
-\citep{visscher2017}. In spite of its advantages, population structure and
+Genome-wide association studies (GWAS)
+identify genetic variants that are statistically
+associated with a specific trait~\citep{uffelmann2021}.
+GWAS has successfully
+identified many loci that are associated with various human diseases and
+traits~\citep{locke2015,ishigaki2022,mahajan2022,yengo2022,mathieson2023},
+and GWAS results are being actively incorporated into clinical
+practice~\citep{visscher2017}.
+Despite these successes, however, there are fundamental issues with
+In spite of its advantages, population structure and
 demographic histories can heavily influence GWAS results \citep{uffelmann2021},

### 2023-11-07 paper.tex
Update

@@
 traits~\citep{locke2015,ishigaki2022,mahajan2022,yengo2022,mathieson2023},
-and GWAS results are being actively incorporated into clinical
+and GWAS results are actively being incorporated into clinical
 practice~\citep{visscher2017}.
-Despite these successes, however, there are fundamental issues with
-In spite of its advantages, population structure and
-demographic histories can heavily influence GWAS results \citep{uffelmann2021},
-so it would be necessary for researchers to implement these factors when
-conducting GWAS.
+Despite these successes and the ongoing collection of population-scale
+Biobank data~\citep{tanjo2021practical}, there are substantial open
+questions around GWAS methology.
+In particular, correctly accounting for population stratification
+remains challenging~\citep{uffelmann2021}.
 
-Ancestral recombination graphs (ARGs) have many promising aspects in GWAS.
+Ancestral recombination graphs (ARGs) describe the interwoven paths
+of genetic inheritance arising from recombination~\citep{wong2023general}.
+[TODO from here]
 Recent studies have shown that genealogy-based association can detect more

### 2023-11-08 paper.tex
Part-way through intro

@@
 %as plain as possible.}
-\keywords{Ancestral Recombination Graph, ARG, GWAS, quantitative traits}
+\keywords{Ancestral Recombination Graph, ARG, GWAS, quantitative traits,
+simulation}
 
@@
 associated with a specific trait~\citep{uffelmann2021}.
-GWAS has successfully
-identified many loci that are associated with various human diseases and
-traits~\citep{locke2015,ishigaki2022,mahajan2022,yengo2022,mathieson2023},
+Many loci that are associated with various human diseases and
+traits have been identified~\citep{locke2015,ishigaki2022,
+mahajan2022,yengo2022,mathieson2023},
 and GWAS results are actively being incorporated into clinical
@@
 In particular, correctly accounting for population stratification
-remains challenging~\citep{uffelmann2021}.
+remain challenging~\citep{uffelmann2021}.
 
-Ancestral recombination graphs (ARGs) describe the interwoven paths
-of genetic inheritance arising from recombination~\citep{wong2023general}.
-[TODO from here]
-Recent studies have shown that genealogy-based association can detect more
-ultra rare variants than conventional methods \citep{zhang2023}, using an
-ARG-based approach in quantitative-trait locus mapping can better detect causal
-locus, especially in the presence of allelic heterogeneity \citep{link2023},
-and linkage disequilibirium can be efficiently modeled in GWAS by using a
-tree-based approach \citep{salehi2023}. These recent advances make ARG to be a
-suitable data format for conducting GWAS.
+Ancestral recombination graphs (ARGs) encode the interwoven paths
+of genetic inheritance arising from recombination~\citep{wong2023general}, and recent
+breakthroughs have opened many new possibilities for their
+practical application~\citep{lewanski2023era}.
+Building on these breakthroughs, substantial progress has been made
+on statistical genetics questions.
+For example,  ARG-based methods
+can detect more ultra rare variants than conventional association testing
+methods~\citep{zhang2023};
+can better detect causal loci in quantitative-trait locus mapping~\citep{link2023tree};
+and can provide a sparse and efficient model of linkage disequilibirium
+in GWAS and downstream applications~\citep{nowbandegani2023extremely}.
 
-ARG is especially useful in population genetic simulations, and it would be
-possible to simulate genetic data with ancestral information by using ARG.
-Simulation is used in GWAS to examine how population structure and demographic
-histories influence polygenic risk score prediction
-\citep{martin2017,zaidi2020}, because simulation is the only way to obtain
-genetic data where the underlying demographic information is known. A succinct
-tree sequence (or tree sequence) is a data structure for ARGs, and it is one of
-the most popular genomic data structures in population genetic simulations
-\citep{baumdicker2022,haller2019}, (cite "A general and efficient, ..."
-article). Since it can efficiently store genetic information with far less data
-size than other methods, it is an appealing choice for simulating mutation and
-ancestry information \citep{kelleher2019}.
+Simulation is a critical component of GWAS method development, and
+generally consists of two steps: (1) generate a synthetic genetic
+variation dataset; and (2) simulate some quantitative traits
+based on this dataset.
+(The two steps may sometimes be combined
+in a single simulation, or we may use real data instead of simulations in
+step (2), but the basic principles are still the same.)
+The combined genetic variation and
+simulated quantitative traits then represent the ``truth'' against
+which association testing methods can be evaluated.
+The first step, of generating realistic, large-scale genetic variation
+datasets, is a well-studied problem with a mature and robust supporting
+software ecosystem. ARG-based methods~\citep{kelleher2016efficient,
+kelleher2018efficient,haller2018tree,baumdicker2022efficient,anderson2023}
+have been instrumental in making simulations of the scale and realism
+required to model modern GWAS cohorts~\cite[e.g.][]{martin2017,zaidi2020}.
+These simulation methods are all based on the ``succinct tree
+sequence'' ARG encoding~\citep{kelleher2019,ralph2020,wong2023general},
+using the open-source \texttt{tskit} library.
+
+% ARG is especially useful in population genetic simulations, and it would be
+% possible to simulate genetic data with ancestral information by using ARG.
+% Simulation is used in GWAS to examine how population structure and demographic
+% histories influence polygenic risk score prediction
+% \citep{martin2017,zaidi2020}, because simulation is the only way to obtain
+% genetic data where the underlying demographic information is known. A succinct
+% tree sequence (or tree sequence) is a data structure for ARGs, and it is one of
+% the most popular genomic data structures in population genetic simulations
+% \citep{baumdicker2022,haller2019}, (cite "A general and efficient, ..."
+% article). Since it can efficiently store genetic information with far less data
+% size than other methods, it is an appealing choice for simulating mutation and
+% ancestry information \citep{kelleher2019}.
 
@@
 population structure. Multiple studies highlight the advantages of using the
-ARG data structure in GWAS \citep{link2023,salehi2023,zhang2023}. ARG data
+ARG data structure in GWAS \citep{link2023tree,nowbandegani2023extremely,zhang2023}.
+ARG data
 structure is becoming increasingly useful in genetic studies, as it is possible

### 2023-11-08 paper.tex
First pass at full intro

@@
 
-% ARG is especially useful in population genetic simulations, and it would be
-% possible to simulate genetic data with ancestral information by using ARG.
-% Simulation is used in GWAS to examine how population structure and demographic
-% histories influence polygenic risk score prediction
-% \citep{martin2017,zaidi2020}, because simulation is the only way to obtain
-% genetic data where the underlying demographic information is known. A succinct
-% tree sequence (or tree sequence) is a data structure for ARGs, and it is one of
-% the most popular genomic data structures in population genetic simulations
-% \citep{baumdicker2022,haller2019}, (cite "A general and efficient, ..."
-% article). Since it can efficiently store genetic information with far less data
-% size than other methods, it is an appealing choice for simulating mutation and
-% ancestry information \citep{kelleher2019}.
+An important property of the succinct tree sequence ARG encoding
+used by these population genetic simulators is that it allows us to
+store vast genetic variation datasets very compactly. Rather than
+storing the genetic sequence variation in its conventional matrix
+form, we can instead store the underlying genealogical history
+and mutations that ultimately \emph{result in} the observed
+variation~\citep{kelleher2019}.
+For example, the ultra-realistic simulations of 1.4 million
+French-Canadian individuals provided by \citet{anderson2023}
+requires around 550 GB of storage space in the
+standard gzip-compressed VCF format ($\sim$280TB uncompressed)
+for chromosome 9. The original simulation file representing
+precisely the same genetic variation data, \emph{plus} the
+complete simulated ARG, requires only 1.36GB of space
+in the compressed \texttt{tszip} format. Thus, in this
+case we achieve more than 400-fold compression,
+and this massive and otherwise highly unwieldy dataset
+becomes straightforward to work with (we discuss
+landmark dataset again in later sections).
+Another substantial benefit of storing simulated datasets
+in this ARG-based format is that many computations can
+be expressed efficiently in terms of the underlying
+ancestry~\citep{kelleher2016efficient,ralph2020}, without
+needing to decode the actual variation data.
 
-There are multiple useful software packages to simulate genotypic data with
-quantitative traits \citep{gaynor2021,haller2023}, and to simulate quantitative
-traits from observed genetic sequences \citep{fernandes2020,meyer2018}.
-However, there does not exist any simulators that can efficiently simulate
-quantitative traits of individuals directly from an ARG. While it would be
-possible to convert a tree sequence data to a different data format and use the
-available packages to simulate quantitative traits, data conversion can be
-computationally costly and it diminishes the computational advantages of using
-tree sequence data. It would be much more efficient to analyze genetic
-information by using the tree sequence data structure, and many population
-genetic statistics can be efficiently computed from the tree structure instead
-of obtaining the genotype matrix \citep{ralph2020}.
+The first step of simulating data for evaluating GWAS methods,
+generating realistic genetic variation datasets of millions
+of individuals, is therefore at a high-level of sophistication
+and more than sufficient to support ongoing research.
+The second step, of generating the actual quantiative traits,
+is currently lagging far behind.
+Although there are multiple packages to simulate genotypic data
+concurrently with quantitative traits~\citep{gaynor2021,haller2023},
+and to simulate quantitative traits from a given set of genetic
+sequences~\citep[e.g.][]{meyer2018,fernandes2020}, none can
+handle the millions of samples now required.
+In particular, there is no software package available to simulate quantitative
+traits directly from the efficient ARG representation used by
+modern population genetic simulators.
 
-In this paper, we propose a new Python library \texttt{tstrait} to simulate
-quantitative traits of individuals in an arbitrary ARG. \texttt{tstrait} is
-built on top of \texttt{tskit} \citep{ralph2020}, which provides an efficient
-way for analyzing and processing ARGs encoded as tree sequence.
-\texttt{tstrait} utilizes the efficient data structure in the tree sequence
-data to simulate complex traits at a fast computational speed. It also provides
-an easily accessible interface by using \texttt{pandas} dataframe
-\citep{pandas} as an input and output of the simulation. Thus, \texttt{tstrait}
-can be easily integrated with other population genetic simulators as well.
-
-In addition to its computational advantages of processing tree sequence data,
-\texttt{tstrait} offers other advantages: \emph{i)} it can simulate traits even
-if the sites have multiple mutations, unlike other simulation frameworks that
-assume infinite-sites model \citep{martin2017}, \emph{ii)} it can simulate
-traits from any ancestral individuals that existed in the demographic history,
-unlike other packages that can only simulate traits from the observed genetic
-information \citep{fernandes2020,meyer2018}, and \emph{iii)} it integrates with
-Python data science packages, such as \texttt{NumPy} \citep{numpy} and
-\texttt{pandas} \citep{pandas}. Due to these advantages, we hope that
-\texttt{tstrait} can provide easily accessible state-of-art quantitative trait
-simulation for researchers who are using population genetic simulation tools.
+In this paper we present \texttt{tstrait}, a new Python library that
+efficiently simulates quantitative traits on an arbitrary ARG.
+The package is build on top of
+\texttt{tskit}~\citep{ralph2020}, which provides a powerful
+framework for storing and processing ARGs.
+Using these efficient data structures, \texttt{tstrait} can
+quickly simulate quantitative traits for population-scale datasets,
+with a very low memory overhead. This approach also provides a
+richer model than existing methods,
+allowing us to, for example, simulate traits for
+ancestral individuals as well as contemporary samples.
+The \texttt{tstrait} library also integrates well with the wider
+PyData ecosystem of data-science tools,
+using
+\texttt{NumPy} arrays~\citep{numpy} and
+\texttt{pandas} dataframes~\citep{pandas}
+as input and output.
 

### 2023-11-08 paper.tex
Updates

@@
 handle the millions of samples now required.
+[TODO discuss bespoke stuff like Martin et al]
 In particular, there is no software package available to simulate quantitative
@@
 
-\section{Phenotypic Model}
+\section{Results}
+\subsection{Phenotypic Model}
+The additive model is a common assumption in GWAS
+\citep{uffelmann2021}, and we assume that a quantitative trait $y$
+is given by
+\[
+y=X\beta+\epsilon,
+\]
+where $\beta$ is the effect size, $X$ is a matrix describing the number
+of causal mutations in the individual, and $\epsilon$ is the environmental
+noise.
+[TODO more discussion here of what this means in an ARG. Be clear about
+causal mutations]
+% Environmental noise is normally distributed with mean zero
+% and a variance of $V_G(1-h^2)/{h^2}$,
+% % distribution,
+% % \begin{align}\label{eq:env}
+%     % \epsilon\sim \mathcal{N}\left(0,\frac{V_G(1-h^2)}{h^2}\right),
+% % \end{align}
+% where $V_G$ is the additive genetic variance from the simulated effect sizes,
+% and $h^2$ is the narrow-sense heritability provided as input by the user.
+% The \texttt{tstrait} API is modular, however, and other forms of environmental
+% noise can be easily generated by the user, if they wish.
 
-Complex traits are traits that are influenced by a mixture of genetic and
-environmental factors. The additive model is a common assumption used in GWAS
-\citep{uffelmann2021}, and \texttt{tstrait} assumes that the traits are
-obtained from the following phenotypic model,
-\begin{align}\label{eq:additive-model}
-    y=X\beta+\epsilon,
-\end{align}
-where $\beta$ is the effect size, $X$ is the matrix that describes the number
-of causal mutations in the individual, and $\epsilon$ is the environmental
-noise. The environmental noise is simulated from the following normal
-distribution,
-\begin{align}\label{eq:env}
-    \epsilon\sim \mathcal{N}\left(0,\frac{V_G(1-h^2)}{h^2}\right),
-\end{align}
-where $V_G$ is the additive genetic variance from the simulated effect sizes
-and $h^2$ is the narrow-sense heritability that is defined by the user.
-
-\section{Package Overview}
-
-\subsection{Effect Size}
-
-The first step of quantitative trait simulation in \texttt{tstrait} is to set
-the trait model distribution by using \emph{trait\_model} function, as in the
-following example:
+\subsection{Effect Sizes}
+The first step of quantitative trait simulation in \texttt{tstrait} is to
+define the trait model distribution, as follows:
 \begin{lstlisting}[language=Python]
 model = tstrait.trait_model(
-  distribution='normal', mean=0, var=1
-)
+    distribution='normal', mean=0, var=1)
 \end{lstlisting}
-The \texttt{tstrait} package supports effect size simulation from five
-univariate distributions and one multivariate distribution, which are taken
-from commonly used statistical distributions to simulate effect sizes
-\citep{gaynor2021,haller2023}. The multivariate normal distribution trait model
-is used to simulate multiple traits, assuming that they are pleiotropic. The
-trait model is used as an input of \emph{sim\_trait} function to simulate
-effect sizes:
+\texttt{Tstrait} currently supports effect size simulation from five
+commonly used univariate distributions~\citep{gaynor2021,haller2023}
+and one multivariate distribution.
+The multivariate normal distribution is used to simulate multiple
+pleiotropic traits.
+The trait model is used as an input of \texttt{sim\_trait} function to simulate
+effect sizes, along with the input ARG and number of causal sites:
 \begin{lstlisting}[language=Python]
 trait_df = tstrait.sim_trait(
-   ts, num_causal=5, model=model
-)
+    arg, num_causal=5, model=model)
 \end{lstlisting}
-\emph{ts} is the tree sequence input, and the user can set the number of causal
-sites that is affecting the trait. The output will be a \texttt{pandas}
-dataframe \citep{pandas} that describes the simulated effect sizes.
+The output, \texttt{trait\_df} is a \texttt{pandas} dataframe
+describing the simulated effect sizes and corresponding
+causal site IDs.
 
 \subsection{Genetic Value}
-
-Genetic value is simulated in \texttt{tstrait} by using the effect size dataframe:
+Given a set of causal sites and effect sizes defined in a dataframe
+and the input ARG, the \emph{genetic values} [TODO DEFINE ME]
+are simulated as follows:
 \begin{lstlisting}[language=Python]
 genetic_result = tstrait.sim_genetic(
-   ts, trait_df, alpha=0
-)
+   arg, trait_df, alpha=0)
 \end{lstlisting}
-\texttt{tstrait} offers flexibility in genetic value simulation, as users can
-define their own effect size dataframe, in addition to using the simulated
-output of \emph{sim\_trait} function. This allows users to use the output from
-other simulators to simulate traits in \texttt{tstrait}.
+The $\alpha$ parameter here is used to
+supports effect size simulation from a frequency
+dependent model, where the simulated effect sizes from the causal mutation
+having allele frequency $p$ are multiplied by $\left(2p(1-p)\right)^\alpha$.
+An $\alpha$ value of zero as in this example means there is
+no frequency dependence. A negative value of $\alpha$
+can be used to model the increased effect sizes on rare variants
+resulting from negative selection~\citep{speed2017}.
+[TODO is this correct? What's the interpretation of positive alpha?]
 
-This is the most computationally intensive part of the simulation algorithm, as
-it would be necessary for the algorithm to count the number of causal mutations
-in each individual inside the dataset. The \texttt{sim\_genetic} function uses
-the tree-traversal algorithm to analyze the tree sequence data
-\citep{ralph2020}, and uses the Python package \texttt{Numba} \citep{numba} to
-speed up the code. This enables \texttt{tstrait} to be an efficient
-quantitative trait simulator of a tree sequence data.
+A major benefit of this modular approach to simulating effect sizes
+and genetic values separately is the flexibility this offers
+users. Because the causal sites and effect sizes are specified in a
+simple tabular format, users can easily develop their own approach
+to simulating these values. Alternatively, other simulators
+such as SLiM~\citep{haller2023} that generate effect sizes and
+causal mutations during the progress of a forwards-time simulation
+(to model the effect of selection, for example) could output these
+values to a CSV or similar file. The modularity here and simple
+input data format is specifically intended to facilitate such
+interoperability between simulators.
 
-\texttt{tstrait} also supports effect size simulation from the frequency
-dependent model, where the simulated effect sizes from the causal mutation
-having allele frequency $p$ will be multiplied by $[2p(1-p)]^\alpha$. The
-$\alpha$ parameter is a user-defined parameter, which is used to control the
-degree of frequency dependence on simulated effect sizes. It has been suggested
-that effect sizes from rarer variants can have increased magnitude from
-negative selection, and the model above with a negative $\alpha$ value is used
-to conduct realistic simulations \citep{speed2017}.
+Generating the genetic values for a trait is by far the
+most computationally intensive part of the simulation,
+as we are essentially counting the number
+of causal mutations in each individual. This is done by a
+straightforward tree traversal algorithm in \texttt{tstrait},
+accelerated using the \texttt{numba} JIT compiler for Python.
+As we see in later sections, this yields very good performance
+at the largest scale, without requiring any additional code in
+compiled languages such as C.
 
 \subsection{Phenotype}
-
-Phenotype is simulated from \texttt{tstrait} by using \emph{sim\_env}:
+The final phenotype simulation is done in \texttt{tstrait}
+by imposing some environmental noise.
+We assume environmental noise is normally distributed with mean zero
+and a variance of
+\[
+V_G(1-h^2)/{h^2}
+\]
+where $V_G$ is the additive genetic variance from the simulated effect sizes,
+and $h^2$ is the narrow-sense heritability parameter provided as input.
+Thus, the \texttt{sim\_env} function takes the dataframe of genetic values
+and narrow-sense heritability as parameters:
 \begin{lstlisting}[language=Python]
 phenotype_df = tstrait.sim_env(
-   genetic_result.genetic, h2=0.3, random_seed=1
-)
+    genetic_result.genetic, h2=0.3)
 \end{lstlisting}
-The parameters are used to simulate environmental noise from the distribution (\ref{eq:env}).
+This simulation returns a dataframe
+describing the simulated genetic value, environmental noise and
+resulting phenotype for each individual at each causal site.
 
-After completing the simulation, \texttt{tstrait} returns two dataframes to
-describe the simulated genetic effect sizes and phenotypes. The resulting
-dataframes include information regarding the site IDs, individual IDs and
-allele frequencies, so it would be possible to regenerate the simulated results
-whenever required.
+As discussed above, there are significant advantages to providing a modular
+API. For convenience, we also provide a single
+\texttt{sim\_phenotype} function that combines the process of simulating
+effect sizes, generating genetic values and imposing environmental noise.
 
-Alternatively, all the above steps of the simulation can be conducted by using
-\emph{sim\_phenotype}:
-\begin{lstlisting}[language=Python]
-sim_result = tstrait.sim_phenotype(
-  ts, num_causal=100, model=model, h2=0.3
-)
-\end{lstlisting}
-The user only needs to input the number of causal sites, trait model and the
-narrow-sense heritability to simulate quantitative traits. By using this
-function, the user does not need to modify the details of the quantitative
-trait simulation.
-
-\section{Computational Time}
-
+\subsection{Performance}
+[START HERE]
 \texttt{tstrait} offers a very efficient way of simulating quantitative traits

### 2023-11-10 paper.tex

@@
 \subsection{Performance}
-[START HERE]
-\texttt{tstrait} offers a very efficient way of simulating quantitative traits
-based on an ARG. Figure \ref{fig:time} (A) shows how the simulation time scales
-with the number of individuals on simulated human genomes in \texttt{stdpopsim}
-\citep{adrion2020}. The only published code to simulate quantitative traits
+\texttt{Tstrait} is very efficient, and can be applied to datasets at the
+largest scales.
+Fig~\ref{fig:time}A shows how the simulation time scales
+with the number of individuals using simple simulations
+of humans using \texttt{stdpopsim}~\citep{adrion2020}.
+[HERE]
+The only published code to simulate quantitative traits
 directly from ARGs is in \cite{martin2017}, which we compared with. Including
@@
 \centering
+% FIXME get the sizes right here and in the notebook where we're generating
+% the figure so we don't have to scale here.
 \includegraphics[width=213pt]{figures/time-scaling.pdf}

### 2023-11-13 paper.tex
Minor edits from Gertjan + Daiki

@@
 in applying ARG-based methods across a broad range of applications.
-Genome-wide association studies (GWAS) are one of the most imporant
+Genome-wide association studies (GWAS) are one of the most important
 application areas, with ARG-based methods already providing substantial
@@
 questions around GWAS methology.
+% TODO rephrase
 In particular, correctly accounting for population stratification
-remain challenging~\citep{uffelmann2021}.
+remains challenging~\citep{uffelmann2021}.
 
@@
 methods~\citep{zhang2023};
-can better detect causal loci in quantitative-trait locus mapping~\citep{link2023tree};
+have better power to detect causal loci in quantitative-trait locus mapping~\citep{link2023tree};
 and can provide a sparse and efficient model of linkage disequilibirium
@@
 sequences~\citep[e.g.][]{meyer2018,fernandes2020}, none can
-handle the millions of samples now required.
+handle the millions of individuals now required.
 [TODO discuss bespoke stuff like Martin et al]
@@
 \subsection{Effect Sizes}
-The first step of quantitative trait simulation in \texttt{tstrait} is to
+The first step of a quantitative trait simulation in \texttt{tstrait} is to
 define the trait model distribution, as follows:

### 2023-11-13 paper.tex

@@
 of individuals, is therefore at a high-level of sophistication
-and more than sufficient to support ongoing research.
+and seems sufficient to support ongoing research.
 The second step, of generating the actual quantiative traits,
@@
 handle the millions of individuals now required.
-[TODO discuss bespoke stuff like Martin et al]
-In particular, there is no software package available to simulate quantitative
+% TODO this needs another pass. We do need to address the Martin et
+% al thing up front though, I think.
+\citet{martin2017} simulated quantitative traits for
+600,000 humans as part of their analysis of the portability
+of polygenic risk scores across populations. This was
+done by using custom scripts to generate traits for these
+large ARGs output by the \texttt{msprime}
+simulator~\cite{kelleher2016efficient}.
+While the code provided by Martin et al.\ was sufficient
+for their analysis, it is tightly coupled to the details of their study
+and not documented or designed to be reused.
+Simulation code can be subtle and difficult to
+debug~\citep{ragsdale2020lessons}, and therefore there
+is a critical need for a well-documented and
+thoroughly tested means of simulating quantitative
 traits directly from the efficient ARG representation used by

### 2023-11-14 paper.tex
Tweaks to intro

@@
 practice~\citep{visscher2017}.
-Despite these successes and the ongoing collection of population-scale
-Biobank data~\citep{tanjo2021practical}, there are substantial open
-questions around GWAS methology.
-% TODO rephrase
-In particular, correctly accounting for population stratification
-remains challenging~\citep{uffelmann2021}.
+This success has led to the collection of many population-scale Biobank
+datasets~\citep{tanjo2021practical}.
+This dense sampling and resulting vast volumes of data, however,
+are presenting significant challenges to GWAS methology.
+For example, how to correctly account for population stratification
+remains an open question~\citep{uffelmann2021}.
 
@@
 of genetic inheritance arising from recombination~\citep{wong2023general}, and recent
-breakthroughs have opened many new possibilities for their
+breakthroughs in inference have opened many new possibilities for their
 practical application~\citep{lewanski2023era}.
-Building on these breakthroughs, substantial progress has been made
-on statistical genetics questions.
+Building on these recent developments, substantial progress has been made
+on statistical genetics applications.
 For example,  ARG-based methods
@@
 methods~\citep{zhang2023};
-have better power to detect causal loci in quantitative-trait locus mapping~\citep{link2023tree};
+have better power to detect causal loci in
+quantitative-trait locus mapping~\citep{link2023tree};
 and can provide a sparse and efficient model of linkage disequilibirium
@@
 software ecosystem. ARG-based methods~\citep{kelleher2016efficient,
-kelleher2018efficient,haller2018tree,baumdicker2022efficient,anderson2023}
-have been instrumental in making simulations of the scale and realism
+kelleher2018efficient,haller2018tree,adrion2020,baumdicker2022efficient,anderson2023}
+have been instrumental in making population genetic simulations
+of the scale and realism
 required to model modern GWAS cohorts~\cite[e.g.][]{martin2017,zaidi2020}.
@@
 used by these population genetic simulators is that it allows us to
-store vast genetic variation datasets very compactly. Rather than
-storing the genetic sequence variation in its conventional matrix
-form, we can instead store the underlying genealogical history
-and mutations that ultimately \emph{result in} the observed
+store vast genetic variation datasets very compactly. Instead of
+storing the genetic sequence variation in the conventional
+matrix-based form, we can store the underlying genealogical history
+and mutations that ultimately \emph{result} in the observed
 variation~\citep{kelleher2019}.
-For example, the ultra-realistic simulations of 1.4 million
-French-Canadian individuals provided by \citet{anderson2023}
-requires around 550 GB of storage space in the
-standard gzip-compressed VCF format ($\sim$280TB uncompressed)
-for chromosome 9. The original simulation file representing
-precisely the same genetic variation data, \emph{plus} the
+For example, a highly realistic simulation
+of chromosome 9 for 1.4 million
+French-Canadian individuals~\citep{anderson2023}
+requires around 550GB of storage space in the
+standard gzip-compressed VCF~\citep{danecek2011}
+format ($\sim$280TB uncompressed).
+The original simulation file representing
+precisely the same genetic variation data, plus the
 complete simulated ARG, requires only 1.36GB of space
-in the compressed \texttt{tszip} format. Thus, in this
-case we achieve more than 400-fold compression,
-and this massive and otherwise highly unwieldy dataset
+in the compressed \texttt{tszip} format. Thus,
+by encoding in terms of the underlying ARG,
+in this case we achieve more than 400-fold compression,
+and this highly unwieldy dataset
 becomes straightforward to work with (we discuss
@@
 In this paper we present \texttt{tstrait}, a new Python library that
-efficiently simulates quantitative traits on an arbitrary ARG.
-The package is build on top of
-\texttt{tskit}~\citep{ralph2020}, which provides a powerful
-framework for storing and processing ARGs.
-Using these efficient data structures, \texttt{tstrait} can
+efficiently simulates quantitative traits on an arbitrary ARG
+in \texttt{tskit}~\citep{ralph2020} format.
+Using the efficient data structures provided by \texttt{tskit},
+\texttt{tstrait} can
 quickly simulate quantitative traits for population-scale datasets,

### 2023-11-14 paper.tex

@@
 used by these population genetic simulators is that it allows us to
-store vast genetic variation datasets very compactly. Instead of
+store enormous genetic variation datasets compactly. Instead of
 storing the genetic sequence variation in the conventional
-matrix-based form, we can store the underlying genealogical history
+matrix-based form~\citep[e.g.\ VCF,][]{danecek2011},
+we can store the underlying genealogical history
 and mutations that ultimately \emph{result} in the observed
@@
 French-Canadian individuals~\citep{anderson2023}
-requires around 550GB of storage space in the
-standard gzip-compressed VCF~\citep{danecek2011}
-format ($\sim$280TB uncompressed).
+requires around 550GB of storage space in
+gzip-compressed VCF ($\sim$280TB uncompressed).
 The original simulation file representing
@@
 complete simulated ARG, requires only 1.36GB of space
-in the compressed \texttt{tszip} format. Thus,
+in compressed \texttt{tszip} format. Thus,
 by encoding in terms of the underlying ARG,
@@
 and this highly unwieldy dataset
-becomes straightforward to work with (we discuss
+becomes straightforward to work with (we discuss this
 landmark dataset again in later sections).
@@
 
-The first step of simulating data for evaluating GWAS methods,
-generating realistic genetic variation datasets of millions
-of individuals, is therefore at a high-level of sophistication
-and seems sufficient to support ongoing research.
-The second step, of generating the actual quantiative traits,
-is currently lagging far behind.
-Although there are multiple packages to simulate genotypic data
-concurrently with quantitative traits~\citep{gaynor2021,haller2023},
-and to simulate quantitative traits from a given set of genetic
-sequences~\citep[e.g.][]{meyer2018,fernandes2020}, none can
-handle the millions of individuals now required.
+There are therefore sophisticated methods available for simulating
+ARGs at the population scale, but there is currently no
+easy way to generate quantiative traits based on such an ARG.
+Existing methods to simulate quantitative traits from a given set of
+genetic sequences in standard formats~\citep[e.g.][]{meyer2018,fernandes2020}
+are overwhelmed by the scale of the data. For example,
+just parsing the 280TB of VCF for the French-Canadian
+dataset mentioned above would take many days of CPU time,
+making quantitative trait simulation by this approach unfeasible.
 % TODO this needs another pass. We do need to address the Martin et
 % al thing up front though, I think.
-\citet{martin2017} simulated quantitative traits for
-600,000 humans as part of their analysis of the portability
-of polygenic risk scores across populations. This was
-done by using custom scripts to generate traits for these
-large ARGs output by the \texttt{msprime}
-simulator~\cite{kelleher2016efficient}.
-While the code provided by Martin et al.\ was sufficient
-for their analysis, it is tightly coupled to the details of their study
-and not documented or designed to be reused.
+In their analysis of the portability of polygenic risk
+scores across populations, \citet{martin2017}
+used efficient \texttt{tskit} APIs to simulate quantitative
+traits for 600,000 humans directly from a simulated ARG.
+While the code provided by Martin et al.\ works at this
+very large scale, it is tightly coupled to the details of their study
+and not designed to be reused.
 Simulation code can be subtle and difficult to
-debug~\citep{ragsdale2020lessons}, and therefore there
-is a critical need for a well-documented and
+debug~\citep{ragsdale2020lessons}, and there
+is therefore a critical need for a well-documented and
 thoroughly tested means of simulating quantitative
-traits directly from the efficient ARG representation used by
-modern population genetic simulators.
+traits directly from an ARG.
 
@@
 in \texttt{tskit}~\citep{ralph2020} format.
-Using the efficient data structures provided by \texttt{tskit},
-\texttt{tstrait} can
+\texttt{Tstrait} can
 quickly simulate quantitative traits for population-scale datasets,
@@
 the Chromosome 9 French Canadian dataset takes approximately 280 terabytes in
-the VCF format, but it is only 1.357 Gb in the \texttt{tskit} compressed
+VCF, but it is only 1.357 Gb in the \texttt{tskit} compressed
 tree-sequence format and it only takes a few minutes to load and process by

### 2023-11-14 paper.tex
Update on performance

@@
 largest scales.
-Fig~\ref{fig:time}A shows how the simulation time scales
-with the number of individuals using simple simulations
-of humans using \texttt{stdpopsim}~\citep{adrion2020}.
+Fig~\ref{fig:time}A shows how the trait simulation time scales
+with number of individuals, for sequences of length 50, 100 and 200Mb.
+For each sample size and sequence length,
+we first simulate an ARG under a simple coalescent model with
+human-like parameters using~\texttt{stdpopsim}~\citep{adrion2020}.
+We then run a quantitative trait simulation
+with 1000 causal sites on this ARG using \texttt{tstrait},
+and report the mean CPU time over 10 independent replicates.
+For comparison, we also report the CPU time required by a
+modified version of the code used by \citet{martin2017} for a
+subset of these simulated ARGs.
+A comparison with existing methods that require reading
+in genetic sequences in formats such as
+VCF~\citep[e.g.][]{meyer2018,fernandes2020} is not feasible,
+as the time required to read the data at this scale would be
+far more than the minutes required here to perform
+the simulations.
+We do not report the memory usage here, but note that these
+simulations for up to 5 million humans were run without difficult
+on a laptop.
+
+For each replicate of the simulation, \texttt{tstrait} is
+used to simulate one quantitative trait with 1000 causal sites. (B)
+
 [HERE]

### 2023-11-14 paper.tex
Updates to perf section

@@
 where $\beta$ is the effect size, $X$ is a matrix describing the number
-of causal mutations in the individual, and $\epsilon$ is the environmental
+of causal alleles in the individual, and $\epsilon$ is the environmental
 noise.
@@
 largest scales.
-Fig~\ref{fig:time}A shows how the trait simulation time scales
+Supplementary Fig~\ref{fig:time}A shows how trait simulation time scales
 with number of individuals, for sequences of length 50, 100 and 200Mb.
@@
 VCF~\citep[e.g.][]{meyer2018,fernandes2020} is not feasible,
-as the time required to read the data at this scale would be
-far more than the minutes required here to perform
-the simulations.
-We do not report the memory usage here, but note that these
-simulations for up to 5 million humans were run without difficult
-on a laptop.
-
-For each replicate of the simulation, \texttt{tstrait} is
-used to simulate one quantitative trait with 1000 causal sites. (B)
+as the VCFs involved are very large, and these tools assume that
+the entire dataset fits in RAM.
+[DISCUSS RAM USAGE]
 
@@
 also used to measure the computational performance.
-
-\begin{figure}[!t]%
-\centering
-% FIXME get the sizes right here and in the notebook where we're generating
-% the figure so we don't have to scale here.
-\includegraphics[width=213pt]{figures/time-scaling.pdf}
-\caption{\textbf{Time taken to simulate quantitative traits.} Each point
-represents the mean time for 10 independent runs under different random seeds.
-The times reported are the total CPU time required to simulate quantitative
-traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
-is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
-$\alpha=-1$. (A) Computational time with increasing number of individuals. The
-whole genome dataset from \emph{Homo sapiens} with various number of
-individuals and chromosome length are simulated by using \texttt{stdpopsim}
-\citep{adrion2020}. For each replicate of the simulation, \texttt{tstrait} is
-used to simulate one quantitative trait with 1000 causal sites. (B)
-Computational time with increasing number of causal sites on two large
-realistic ARGs. \texttt{tstrait} is used to simulate quantitative traits from
-the simulated French Canadian dataset \citep{anderson2023} and the inferred
-tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with
-varying number of causal sites. The French Canadian dataset is downloaded from
-\url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000
-Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}.
-Chromosome 9 is selected for both datasets.}\label{fig:time}
-\end{figure}
 
@@
 \section{Funding}
-
-D.T. is supported by the Oxford Kobe Scholarship from the University of Oxford
+DT is supported by the Oxford Kobe Scholarship from the University of Oxford
 and the Euretta J. Kellett Fellowship from Columbia University.
-
-This work was supported by ...
+JK acknowledges support from the Robertson Foundation,
+NIH (research grants HG011395 and HG012473) and
+EPSRC (research grant EP/X024881/1).
 
@@
 
+\newpage
+\renewcommand\thefigure{S\arabic{figure}}
+\setcounter{figure}{0}
+\renewcommand\thetable{S\arabic{table}}
+\setcounter{table}{0}
+\section{Supplementary Figures}
+
+\begin{figure*}[t]%
+\centering
+% FIXME get the sizes right here and in the notebook where we're generating
+% the figure so we don't have to scale here.
+\includegraphics[width=213pt]{figures/time-scaling.pdf}
+\caption{\textbf{Time taken to simulate quantitative traits.} Each point
+represents the mean time for 10 independent runs under different random seeds.
+The times reported are the total CPU time required to simulate quantitative
+traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
+is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
+$\alpha=-1$. (A) CPU time with increasing number of individuals
+based on ARGs simulated using \texttt{stdpopsim} with the [MODEL ID]
+demographic mode. For each replicate we simulate
+one quantitative trait with 1000 causal sites.
+(B) CPU time with increasing number of causal sites on two large ARGs.
+\texttt{tstrait} is used to simulate quantitative traits from
+the simulated French Canadian dataset \citep{anderson2023} and the inferred
+tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with
+varying number of causal sites. The French Canadian dataset is downloaded from
+\url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000
+Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}.
+Chromosome 9 is selected for both datasets.}\label{fig:time}
+\end{figure*}
+
 \end{document}

### 2023-11-15 paper.tex
Update caption

@@
 
+\begin{figure}[t]%
+\centering
+\includegraphics[width=240pt]{figures/tree-illustration.pdf}
+\caption{Example simulation of a phenotype at a site with ancestral state A and
+two mutations. In this diploid example each of the three individuals
+is assocated with two nodes (i.e., the individual with ID 0 corresponds
+to nodes \textsf{a} and \textsf{b}). Internal nodes in the tree are associated
+with the null individual, $-1$. Here, the trait's causal allele is T
+and an effect size $\beta=0.05$. Each node in the tree has an associated
+genetic value, and the overall genetic value for an individual is the
+sum of their corresponding node values.
+\label{fig:tree-illustration}}
+\end{figure}
+
 \subsection{Genetic Value}
@@
 
-\newpage
+\clearpage
+
 \renewcommand\thefigure{S\arabic{figure}}

### 2023-11-15 paper.tex
WIP

@@
 \section{Results}
-\subsection{Phenotypic Model}
-The additive model is a common assumption in GWAS
-\citep{uffelmann2021}, and we assume that a quantitative trait $y$
+\begin{figure}[t]%
+\centering
+\includegraphics[width=240pt]{figures/tree-illustration.pdf}
+\caption{Example simulation of a phenotype at a site with ancestral state A and
+two mutations. In this diploid example each of the three individuals
+is assocated with two nodes (i.e., the individual with ID 0 corresponds
+to nodes \textsf{a} and \textsf{b}). Internal nodes in the tree are associated
+with the null individual, $-1$. Here, the trait's causal allele is T
+and effect size $\beta=0.05$. Each node in the tree has an associated
+genetic value, and the overall genetic value for an individual is the
+sum of their corresponding node values. The final phenotype
+for each individual is the sum of the genetic value and
+simulated environmental noise.
+\label{fig:tree-illustration}}
+\end{figure}
+
+\subsection{Model}
+Phenotypes are simulated in \texttt{tstrait} following standard
+models, adapted to the ARG context (Fig~\ref{fig:tree-illustration}).
+Each trait is associated with one or more causal sites (position
+on the genome), and at each causal site there is a causal allele
+(i.e.\ a particular nucleotide). Because we are reasoning
+about the inheritance of particular alleles at a given position,
+we can consider just the local tree at that site and
+the network structure of the ARG along the genome is not
+important~\citep{wong2023general}. In
+Fig.~\ref{fig:tree-illustration} we show the local tree
+relating three diploid individuals for a given causal site.
+Because these individuals are diploids, each is associated
+with two nodes in the tree (highlighted by colour). Ancestral
+nodes are not associated with individuals.
+
+[WIP]
+At a given site, the causal allele is associated with an
+effect size $\beta$,
+and every node in the tree that carries this allele has a
+``genetic value'' of $\beta$.
+
+Following the standard practise in GWAS~\citep{uffelmann2021}
+we assume the additive model, and a quantitative trait $y$
 is given by
@@
 noise.
-[TODO more discussion here of what this means in an ARG. Be clear about
-causal mutations]
-% Environmental noise is normally distributed with mean zero
-% and a variance of $V_G(1-h^2)/{h^2}$,
-% % distribution,
-% % \begin{align}\label{eq:env}
-%     % \epsilon\sim \mathcal{N}\left(0,\frac{V_G(1-h^2)}{h^2}\right),
-% % \end{align}
-% where $V_G$ is the additive genetic variance from the simulated effect sizes,
-% and $h^2$ is the narrow-sense heritability provided as input by the user.
-% The \texttt{tstrait} API is modular, however, and other forms of environmental
-% noise can be easily generated by the user, if they wish.
+
+Environmental noise is normally distributed with mean zero
+and a variance of $V_G(1-h^2)/{h^2}$,
+% distribution,
+% \begin{align}\label{eq:env}
+    % \epsilon\sim \mathcal{N}\left(0,\frac{V_G(1-h^2)}{h^2}\right),
+% \end{align}
+where $V_G$ is the additive genetic variance from the simulated effect sizes,
+and $h^2$ is the narrow-sense heritability provided as input by the user.
+The \texttt{tstrait} API is modular, however, and other forms of environmental
+noise can be easily generated by the user, if they wish.
 
@@
 causal site IDs.
-
-\begin{figure}[t]%
-\centering
-\includegraphics[width=240pt]{figures/tree-illustration.pdf}
-\caption{Example simulation of a phenotype at a site with ancestral state A and
-two mutations. In this diploid example each of the three individuals
-is assocated with two nodes (i.e., the individual with ID 0 corresponds
-to nodes \textsf{a} and \textsf{b}). Internal nodes in the tree are associated
-with the null individual, $-1$. Here, the trait's causal allele is T
-and an effect size $\beta=0.05$. Each node in the tree has an associated
-genetic value, and the overall genetic value for an individual is the
-sum of their corresponding node values.
-\label{fig:tree-illustration}}
-\end{figure}
 

### 2023-11-16 paper.tex
Pass through what we do

@@
 Ancestral recombination graphs (ARGs) encode the interwoven paths
-of genetic inheritance arising from recombination~\citep{wong2023general}, and recent
+of genetic inheritance arising from
+% Explanation of citations: people might get offended if we don't cite
+% Griffiths when first mentioning the ARG, but Hudson has just as
+% much to do with the actual definition that we use in Wong 2023.
+recombination~\citep{hudson1983properties,griffiths1997ancestral,wong2023general},
+and recent
 breakthroughs in inference have opened many new possibilities for their
@@
 These simulation methods are all based on the ``succinct tree
-sequence'' ARG encoding~\citep{kelleher2019,ralph2020,wong2023general},
+sequence'' ARG encoding~\citep{ralph2020,wong2023general},
 using the open-source \texttt{tskit} library.
@@
 ARGs at the population scale, but there is currently no
-easy way to generate quantiative traits based on such an ARG.
+easy way to generate quantitative traits based on such an ARG.
 Existing methods to simulate quantitative traits from a given set of
@@
 The \texttt{tstrait} library also integrates well with the wider
-PyData ecosystem of data-science tools,
-using
-\texttt{NumPy} arrays~\citep{numpy} and
-\texttt{pandas} dataframes~\citep{pandas}
-as input and output.
+Python data-science ecosytem, allowing users to efficiently analyse
+large-scale data using familiar and ergonomic tools.
 
@@
 (i.e.\ a particular nucleotide). Because we are reasoning
-about the inheritance of particular alleles at a given position,
+about the inheritance of alleles at a site on the genome
 we can consider just the local tree at that site and
-the network structure of the ARG along the genome is not
-important~\citep{wong2023general}. In
-Fig.~\ref{fig:tree-illustration} we show the local tree
+the network structure of the ARG is not important~\citep{wong2023general}.
+In Fig.~\ref{fig:tree-illustration} we show the local tree
 relating three diploid individuals for a given causal site.
 Because these individuals are diploids, each is associated
-with two nodes in the tree (highlighted by colour). Ancestral
-nodes are not associated with individuals.
+with two nodes in the tree (highlighted by colour).
+Ancestral nodes are not associated with individuals here,
+but in general the ARG may be embedded in a multigenerational
+pedigree, where some internal nodes would be associated
+with individuals.
 
-[WIP]
 At a given site, the causal allele is associated with an
@@
 and every node in the tree that carries this allele has a
-``genetic value'' of $\beta$.
+``genetic value'' of $\beta=0.05$. In the example of
+Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
+allele, and therefore all nodes descending from \textsf{i}
+have genetic value $0.05$, except \textsf{e} which
+has zero because of the back-mutation to A.
+Following the standard practise in GWAS~\citep{uffelmann2021}
+we assume the additive model such that the overall
+genetic value of an individual is the sum of its
+nodes' genetic values.
+Given these per-individual genetic values, the final phenotype
+is then generated by adding some environmental noise.
+This noise is normally distributed with mean zero
+and variance of $V_G(1-h^2)/{h^2}$,
+% TODO I find this clearer than saying "additive genetic variance"
+where $V_G$ is the variance of the individual genetic values
+and $h^2$ is the narrow-sense heritability provided as input by the user.
+% The \texttt{tstrait} API is modular, however, and other forms of environmental
+% noise can be easily generated by the user, if they wish.
 
-Following the standard practise in GWAS~\citep{uffelmann2021}
-we assume the additive model, and a quantitative trait $y$
-is given by
-\[
-y=X\beta+\epsilon,
-\]
-where $\beta$ is the effect size, $X$ is a matrix describing the number
-of causal alleles in the individual, and $\epsilon$ is the environmental
-noise.
+\subsection{Interface}
+\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
+ARG library~\citep{ralph2020,wong2023general} and the rich
+Python data-science ecosystem~\citep{numpy}.
+In the simplest case, simulating a phenotype for an ARG requires
+very little code:
+% NOTE this listing assumes num_causal has been moved, or removed.
+% We need to check whether this actually works in the final version.
+\begin{lstlisting}[language=Python]
+import tstrait as tst
+model = tst.trait_model('normal', mean=0, var=1)
+result = tst.sim_phenotype(arg, model, h2=0.3)
+\end{lstlisting}
+We first create \texttt{model} representing the distribution
+from which effect sizes are drawn. Five commonly used
+univariate distributions are supported, along with the
+multivariate normal distribution to model pleiotropic traits.
+Given this model, we can then simulate phenotypes for the individuals
+in an ARG (as a  \texttt{tskit TreeSequence}) using the
+\texttt{sim\_phenotype} function, which also takes the trait model
+and narrow sense heritability $h^2$ as parameters.
+The return value \texttt{result} is an object encapsulating
+two Pandas dataframes~\citep{pandas}: one describing the simulated
+effect sizes and the other describing the genetic values,
+environmental noise and final phenotypes for each individual.
+The simulation results can then be efficiently and conventiently
+processed by a host of Python data science tools.
 
-Environmental noise is normally distributed with mean zero
-and a variance of $V_G(1-h^2)/{h^2}$,
-% distribution,
-% \begin{align}\label{eq:env}
-    % \epsilon\sim \mathcal{N}\left(0,\frac{V_G(1-h^2)}{h^2}\right),
-% \end{align}
-where $V_G$ is the additive genetic variance from the simulated effect sizes,
-and $h^2$ is the narrow-sense heritability provided as input by the user.
-The \texttt{tstrait} API is modular, however, and other forms of environmental
-noise can be easily generated by the user, if they wish.
-
-\subsection{Effect Sizes}
-The first step of a quantitative trait simulation in \texttt{tstrait} is to
-define the trait model distribution, as follows:
-\begin{lstlisting}[language=Python]
-model = tstrait.trait_model(
-    distribution='normal', mean=0, var=1)
-\end{lstlisting}
-\texttt{Tstrait} currently supports effect size simulation from five
-commonly used univariate distributions~\citep{gaynor2021,haller2023}
-and one multivariate distribution.
-The multivariate normal distribution is used to simulate multiple
-pleiotropic traits.
-The trait model is used as an input of \texttt{sim\_trait} function to simulate
-effect sizes, along with the input ARG and number of causal sites:
-\begin{lstlisting}[language=Python]
-trait_df = tstrait.sim_trait(
-    arg, num_causal=5, model=model)
-\end{lstlisting}
-The output, \texttt{trait\_df} is a \texttt{pandas} dataframe
-describing the simulated effect sizes and corresponding
-causal site IDs.
-
-\subsection{Genetic Value}
-Given a set of causal sites and effect sizes defined in a dataframe
-and the input ARG, the \emph{genetic values} [TODO DEFINE ME]
-are simulated as follows:
-\begin{lstlisting}[language=Python]
-genetic_result = tstrait.sim_genetic(
-   arg, trait_df, alpha=0)
-\end{lstlisting}
-The $\alpha$ parameter here is used to
-supports effect size simulation from a frequency
-dependent model, where the simulated effect sizes from the causal mutation
-having allele frequency $p$ are multiplied by $\left(2p(1-p)\right)^\alpha$.
-An $\alpha$ value of zero as in this example means there is
-no frequency dependence. A negative value of $\alpha$
+This simple usage is not very flexible, however,
+and so \texttt{tstrait} provides more modular building blocks
+for power-users, and to potentially integrate with other tools.
+Effect sizes can be simulated for a given ARG and trait model
+using the \texttt{sim\_trait} function.
+This function also supports a form
+of frequency dependence, where the a simulated effect size
+for a causal allele with frequency $p$
+is multiplied by $\left(2p(1-p)\right)^\alpha$.
+A negative value of $\alpha$
 can be used to model the increased effect sizes on rare variants
 resulting from negative selection~\citep{speed2017}.
-[TODO is this correct? What's the interpretation of positive alpha?]
+The return value of the \texttt{sim\_trait} function is a
+data frame describing the simulated effect sizes and causal alleles.
+This dataframe can then be passed to the \texttt{genetic\_values}
+function, which calculates the genetic values for each
+node, and accumulates these per individual
+(Fig~\ref{fig:tree-illustration}). Finally, the \texttt{sim\_env}
+function takes these per-individual genetic values
+and adds some simulated environmental noise to produce the final
+phenotypes (as discussed above).
 
@@
 
-Generating the genetic values for a trait is by far the
-most computationally intensive part of the simulation,
-as we are essentially counting the number
-of causal mutations in each individual. This is done by a
-straightforward tree traversal algorithm in \texttt{tstrait},
-accelerated using the \texttt{numba} JIT compiler for Python.
-As we see in later sections, this yields very good performance
-at the largest scale, without requiring any additional code in
-compiled languages such as C.
+% \subsection{Phenotype}
+% The final phenotype simulation is done in \texttt{tstrait}
+% by imposing some environmental noise.
+% We assume environmental noise is normally distributed with mean zero
+% and a variance of
+% \[
+% V_G(1-h^2)/{h^2}
+% \]
+% where $V_G$ is the additive genetic variance from the simulated effect sizes,
+% and $h^2$ is the narrow-sense heritability parameter provided as input.
+% Thus, the \texttt{sim\_env} function takes the dataframe of genetic values
+% and narrow-sense heritability as parameters:
+% \begin{lstlisting}[language=Python]
+% phenotype_df = tstrait.sim_env(
+%     genetic_result.genetic, h2=0.3)
+% \end{lstlisting}
+% This simulation returns a dataframe
+% describing the simulated genetic value, environmental noise and
+% resulting phenotype for each individual at each causal site.
 
-\subsection{Phenotype}
-The final phenotype simulation is done in \texttt{tstrait}
-by imposing some environmental noise.
-We assume environmental noise is normally distributed with mean zero
-and a variance of
-\[
-V_G(1-h^2)/{h^2}
-\]
-where $V_G$ is the additive genetic variance from the simulated effect sizes,
-and $h^2$ is the narrow-sense heritability parameter provided as input.
-Thus, the \texttt{sim\_env} function takes the dataframe of genetic values
-and narrow-sense heritability as parameters:
-\begin{lstlisting}[language=Python]
-phenotype_df = tstrait.sim_env(
-    genetic_result.genetic, h2=0.3)
-\end{lstlisting}
-This simulation returns a dataframe
-describing the simulated genetic value, environmental noise and
-resulting phenotype for each individual at each causal site.
+% As discussed above, there are significant advantages to providing a modular
+% API. For convenience, we also provide a single
+% \texttt{sim\_phenotype} function that combines the process of simulating
+% effect sizes, generating genetic values and imposing environmental noise.
 
-As discussed above, there are significant advantages to providing a modular
-API. For convenience, we also provide a single
-\texttt{sim\_phenotype} function that combines the process of simulating
-effect sizes, generating genetic values and imposing environmental noise.
+\subsection{Implementation}
+\texttt{Tstrait} is written entirely in Python. Many numerical operations are
+peformed using standard array-oriented approaches and therefore
+highly efficient~\citep{numpy}. Generating the genetic values for a given
+trait is computationally intensive. This is done by a
+straightforward tree traversal algorithm, which is
+accelerated using the \texttt{numba} JIT compiler for Python yielding
+excellent performance.
 

### 2023-11-16 paper.tex
Trim acks

@@
 
-\subsection{Implementation}
+\subsection{Implementation and validation}
 \texttt{Tstrait} is written entirely in Python. Many numerical operations are
@@
 excellent performance.
+
+The \texttt{tstrait} codebase includes an extensive suite of unit tests,
+which are automatically run as part of the development process. The
+output of \texttt{tstrait} has been statistically validated against
+[FILL ME IN].
 
@@
 the entire dataset fits in RAM.
-[DISCUSS RAM USAGE]
-
-[HERE]
-The only published code to simulate quantitative traits
-directly from ARGs is in \cite{martin2017}, which we compared with. Including
-other tools that require the entire genome matrix is not practical, as the
-simulation time will be dominated by processing the VCF file and it would not
-be practically to analyze a large sample VCF file on a laptop PC. For example,
-the Chromosome 9 French Canadian dataset takes approximately 280 terabytes in
-VCF, but it is only 1.357 Gb in the \texttt{tskit} compressed
-tree-sequence format and it only takes a few minutes to load and process by
-using the tree sequence data structure.
+Supplementary Fig~\ref{fig:time}B shows that RAM usage is very
+modest for all these simulations.
 
@@
 \section{Conclusion}
-
 We believe that in the coming years, simulation in tree sequence encoding will
@@
 \section{Acknowledgments}
-
-We greatly acknowledge helpful discussions about quantitative trait simulation
-with Gregor Gorjanc. We thank Ben Haller and people in the tskit community for
-helpful discussions and feedback. We thank Ben Jeffery for his help with the
-Python package development.
+We are grateful to Gregor Gorjanc, Ben Haller, Ben Jeffery and
+others in the tskit community for helpful discussions and feedback.
 

### 2023-11-17 paper.tex
Updates from Daiki

@@
 This success has led to the collection of many population-scale Biobank
-datasets~\citep{tanjo2021practical}.
-This dense sampling and resulting vast volumes of data, however,
-are presenting significant challenges to GWAS methology.
-For example, how to correctly account for population stratification
-remains an open question~\citep{uffelmann2021}.
+datasets~\citep{tanjo2021practical}, 
+but the sampling and resulting vast volumes of data
+are presenting significant challenges to GWAS 
+methology~\citep{uffelmann2021}.
 
@@
 generally consists of two steps: (1) generate a synthetic genetic
-variation dataset; and (2) simulate some quantitative traits
+variation dataset; and (2) simulate quantitative traits
 based on this dataset.
@@
 These simulation methods are all based on the ``succinct tree
-sequence'' ARG encoding~\citep{ralph2020,wong2023general},
-using the open-source \texttt{tskit} library.
+sequence'' ARG encoding~\citep{wong2023general},
+using the open-source \texttt{tskit} library~\citep{ralph2020}.
 
@@
 ARGs at the population scale, but there is currently no
-easy way to generate quantitative traits based on such an ARG.
+easy way to simulate quantitative traits based on such an ARG.
 Existing methods to simulate quantitative traits from a given set of
@@
 debug~\citep{ragsdale2020lessons}, and there
-is therefore a critical need for a well-documented and
+is a critical need for a well-documented and
 thoroughly tested means of simulating quantitative
@@
 efficiently simulates quantitative traits on an arbitrary ARG
-in \texttt{tskit}~\citep{ralph2020} format.
+in \texttt{tskit} format.
 \texttt{Tstrait} can
@@
 The \texttt{tstrait} library also integrates well with the wider
-Python data-science ecosytem, allowing users to efficiently analyse
+Python data-science ecosystem~\citep{numpy}, 
+allowing users to efficiently analyse
 large-scale data using familiar and ergonomic tools.
@@
 two mutations. In this diploid example each of the three individuals
-is assocated with two nodes (i.e., the individual with ID 0 corresponds
+is associated with two nodes (i.e., the individual with ID 0 corresponds
 to nodes \textsf{a} and \textsf{b}). Internal nodes in the tree are associated
@@
 Phenotypes are simulated in \texttt{tstrait} following standard
-models, adapted to the ARG context (Fig~\ref{fig:tree-illustration}).
+models~\citep{uffelmann2021}, adapted to the ARG context (Fig~\ref{fig:tree-illustration}).
 Each trait is associated with one or more causal sites (position
@@
 (i.e.\ a particular nucleotide). Because we are reasoning
-about the inheritance of alleles at a site on the genome
+about the inheritance of alleles at a site on the genome,
 we can consider just the local tree at that site and
@@
 has zero because of the back-mutation to A.
-Following the standard practise in GWAS~\citep{uffelmann2021}
+Following the standard practise in GWAS~\citep{uffelmann2021},
 we assume the additive model such that the overall
@@
 is then generated by adding some environmental noise.
-This noise is normally distributed with mean zero
+This noise is simulated from a normal distribution with mean zero
 and variance of $V_G(1-h^2)/{h^2}$,
@@
 environmental noise and final phenotypes for each individual.
-The simulation results can then be efficiently and conventiently
+The simulation results can then be efficiently and conveniently
 processed by a host of Python data science tools.
@@
 
+% TODO this is to be rewritten with just a single example from FC
+% dataset and optionally also mentioning real data from 1000G (space
+% permitting).
 Figure \ref{fig:time} (B) shows simulation time of \texttt{tstrait} with more

### 2023-11-17 paper.tex
numba citation

@@
 straightforward tree traversal algorithm, which is
-accelerated using the \texttt{numba} JIT compiler for Python yielding
+accelerated using the \texttt{numba} JIT compiler for Python~\citep{numba} yielding
 excellent performance.

### 2023-11-17 paper.tex
switch pandas citation

@@
 The return value \texttt{result} is an object encapsulating
-two Pandas dataframes~\citep{pandas}: one describing the simulated
+two Pandas dataframes~\citep{mckinney2010data}: one describing the simulated
 effect sizes and the other describing the genetic values,

### 2023-11-20 paper.tex
Minor updates

@@
 be expressed efficiently in terms of the underlying
-ancestry~\citep{kelleher2016efficient,ralph2020}, without
+genealogies~\citep{kelleher2016efficient,ralph2020}, without
 needing to decode the actual variation data.
 
-There are therefore sophisticated methods available for simulating
-ARGs at the population scale, but there is currently no
+Although there are sophisticated methods available for simulating
+ARGs at the population scale, there is currently no
 easy way to simulate quantitative traits based on such an ARG.
@@
 making quantitative trait simulation by this approach unfeasible.
-% TODO this needs another pass. We do need to address the Martin et
-% al thing up front though, I think.
 In their analysis of the portability of polygenic risk
 scores across populations, \citet{martin2017}
-used efficient \texttt{tskit} APIs to simulate quantitative
-traits for 600,000 humans directly from a simulated ARG.
-While the code provided by Martin et al.\ works at this
-very large scale, it is tightly coupled to the details of their study
+demonstrated the ability to simulate phenotypes at scale using ARGs. 
+However, their approach is tightly coupled to the details 
+of the study 
 and not designed to be reused.
@@
 ARG data
-structure is becoming increasingly useful in genetic studies, as it is possible
+structures are becoming increasingly useful in genetic studies, as it is possible
 to accurately infer biobank-scale genealogies from sequencing data

### 2023-11-20 paper.tex
Tweaks

@@
 Phenotypes are simulated in \texttt{tstrait} following standard
-models~\citep{uffelmann2021}, adapted to the ARG context (Fig~\ref{fig:tree-illustration}).
-Each trait is associated with one or more causal sites (position
-on the genome), and at each causal site there is a causal allele
-(i.e.\ a particular nucleotide). Because we are reasoning
+models~\citep{uffelmann2021}, adapted to the ARG context.
+Each trait is associated with one or more causal sites
+(positions on the genome), and at each causal site there is a causal allele
+(i.e.\ a particular nucleotide) associated with an effect size $\beta$.
+Every node in the local tree that inherits the causal allele
+at that site is said to have a ``genetic value'' of $\beta$.
+Because we are reasoning
 about the inheritance of alleles at a site on the genome,
@@
 the network structure of the ARG is not important~\citep{wong2023general}.
-In Fig.~\ref{fig:tree-illustration} we show the local tree
-relating three diploid individuals for a given causal site.
+
+In Fig.~\ref{fig:tree-illustration} we show an example
+tree for three individuals.
 Because these individuals are diploids, each is associated
@@
 with individuals.
-
-At a given site, the causal allele is associated with an
-effect size $\beta$,
-and every node in the tree that carries this allele has a
-``genetic value'' of $\beta=0.05$. In the example of
+In the example of
 Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
-allele, and therefore all nodes descending from \textsf{i}
+allele and $\beta=0.05$, and
+therefore all nodes descending from \textsf{i}
 have genetic value $0.05$, except \textsf{e} which
@@
 model = tst.trait_model('normal', mean=0, var=1)
-result = tst.sim_phenotype(arg, model, h2=0.3)
+result = tst.sim_phenotype(arg, model)
 \end{lstlisting}
-We first create \texttt{model} representing the distribution
+We first create \texttt{model}, which represents the distribution
 from which effect sizes are drawn. Five commonly used
@@
 in an ARG (as a  \texttt{tskit TreeSequence}) using the
-\texttt{sim\_phenotype} function, which also takes the trait model
-and narrow sense heritability $h^2$ as parameters.
+\texttt{sim\_phenotype} function.
 The return value \texttt{result} is an object encapsulating
@@
 processed by a host of Python data science tools.
+
+[WIP]
 

### 2023-11-21 paper.tex
Move freq-dep to model section

@@
 Phenotypes are simulated in \texttt{tstrait} following standard
-models~\citep{uffelmann2021}, adapted to the ARG context.
+GWAS models~\citep{uffelmann2021}, adapted to the ARG context.
 Each trait is associated with one or more causal sites
@@
 (i.e.\ a particular nucleotide) associated with an effect size $\beta$.
-Every node in the local tree that inherits the causal allele
-at that site is said to have a ``genetic value'' of $\beta$.
-Because we are reasoning
-about the inheritance of alleles at a site on the genome,
-we can consider just the local tree at that site and
-the network structure of the ARG is not important~\citep{wong2023general}.
+For each causal site, an effect size is drawn from a distribution
+and optionally multiplied by $\left(2p(1-p)\right)^\alpha$,
+where $p$ is the frequency of the causal allele and $\alpha$
+is a parameter describing the strength of
+frequency dependence~\citep{speed2017}.
+At a particular causal site, every node in the local tree that inherits
+the causal allele at that site is said to have a ``genetic value'' of $\beta$.
+% NOTE removing this for space - it keeps this section on one page.
+% Because we are reasoning
+% about the inheritance of alleles at a site on the genome,
+% we can consider just the local tree at that site and
+% the network structure of the ARG is not important~\citep{wong2023general}.
 
@@
 and variance of $V_G(1-h^2)/{h^2}$,
-% TODO I find this clearer than saying "additive genetic variance"
 where $V_G$ is the variance of the individual genetic values
 and $h^2$ is the narrow-sense heritability provided as input by the user.
-% The \texttt{tstrait} API is modular, however, and other forms of environmental
-% noise can be easily generated by the user, if they wish.
 
@@
 Python data-science ecosystem~\citep{numpy}.
-In the simplest case, simulating a phenotype for an ARG requires
-very little code:
+Simulating a phenotype for an ARG with default parameter
+values requires only a few lines of code:
 % NOTE this listing assumes num_causal has been moved, or removed.
@@
 \texttt{sim\_phenotype} function.
+
 The return value \texttt{result} is an object encapsulating
@@
 using the \texttt{sim\_trait} function.
-This function also supports a form
-of frequency dependence, where the a simulated effect size
-for a causal allele with frequency $p$
-is multiplied by $\left(2p(1-p)\right)^\alpha$.
-A negative value of $\alpha$
-can be used to model the increased effect sizes on rare variants
-resulting from negative selection~\citep{speed2017}.
-The return value of the \texttt{sim\_trait} function is a
-data frame describing the simulated effect sizes and causal alleles.
 This dataframe can then be passed to the \texttt{genetic\_values}

### 2023-11-21 paper.tex
Interface

@@
 \texttt{sim\_phenotype} function.
-
 The return value \texttt{result} is an object encapsulating
@@
 The simulation results can then be efficiently and conveniently
-processed by a host of Python data science tools.
+processed by a range of familiar Python data science tools.
 
-[WIP]
-
-This simple usage is not very flexible, however,
-and so \texttt{tstrait} provides more modular building blocks
-for power-users, and to potentially integrate with other tools.
-Effect sizes can be simulated for a given ARG and trait model
-using the \texttt{sim\_trait} function.
+As well as this convenient single-function interface,
+\texttt{tstrait} provides modular building blocks for power-users
+and potential integration with other tools that generate
+traits on an ARG. The \texttt{sim\_trait} function
+simulates effect sizes for an input ARG, and returns a data frame
+describing the causal alleles and effect sizes.
 This dataframe can then be passed to the \texttt{genetic\_values}
 function, which calculates the genetic values for each
-node, and accumulates these per individual
+node, and accumulates them by individual
 (Fig~\ref{fig:tree-illustration}). Finally, the \texttt{sim\_env}
@@
 and adds some simulated environmental noise to produce the final
-phenotypes (as discussed above).
+phenotypes.
 
-A major benefit of this modular approach to simulating effect sizes
-and genetic values separately is the flexibility this offers
+A major benefit of this modular architecture is the flexibility it offers
 users. Because the causal sites and effect sizes are specified in a
@@
 (to model the effect of selection, for example) could output these
-values to a CSV or similar file. The modularity here and simple
-input data format is specifically intended to facilitate such
-interoperability between simulators.
+values to a CSV or similar file. The modular architecture  and simple
+input data formats are specifically intended to facilitate such
+interoperability.
 

### 2023-11-21 paper.tex
Update perf

@@
 
-% \subsection{Phenotype}
-% The final phenotype simulation is done in \texttt{tstrait}
-% by imposing some environmental noise.
-% We assume environmental noise is normally distributed with mean zero
-% and a variance of
-% \[
-% V_G(1-h^2)/{h^2}
-% \]
-% where $V_G$ is the additive genetic variance from the simulated effect sizes,
-% and $h^2$ is the narrow-sense heritability parameter provided as input.
-% Thus, the \texttt{sim\_env} function takes the dataframe of genetic values
-% and narrow-sense heritability as parameters:
-% \begin{lstlisting}[language=Python]
-% phenotype_df = tstrait.sim_env(
-%     genetic_result.genetic, h2=0.3)
-% \end{lstlisting}
-% This simulation returns a dataframe
-% describing the simulated genetic value, environmental noise and
-% resulting phenotype for each individual at each causal site.
-
-% As discussed above, there are significant advantages to providing a modular
-% API. For convenience, we also provide a single
-% \texttt{sim\_phenotype} function that combines the process of simulating
-% effect sizes, generating genetic values and imposing environmental noise.
-
 \subsection{Implementation and validation}
-\texttt{Tstrait} is written entirely in Python. Many numerical operations are
-peformed using standard array-oriented approaches and therefore
-highly efficient~\citep{numpy}. Generating the genetic values for a given
-trait is computationally intensive. This is done by a
-straightforward tree traversal algorithm, which is
-accelerated using the \texttt{numba} JIT compiler for Python~\citep{numba} yielding
-excellent performance.
-
+\texttt{Tstrait} is written entirely in Python. Numerical operations
+are either peformed using standard array-oriented
+approaches~\citep{numpy} or accelerated
+using the \texttt{numba} JIT compiler~\citep{numba}.
 The \texttt{tstrait} codebase includes an extensive suite of unit tests,
 which are automatically run as part of the development process. The
-output of \texttt{tstrait} has been statistically validated against
-[FILL ME IN].
+output of \texttt{tstrait} has been extensively validated against
+theoretical expectations, as well as the output
+of \texttt{AlphaSimR}~\citep{gaynor2021} and
+\texttt{simplePHENOTYPES}~\citep{fernandes2020}.
 
@@
 and report the mean CPU time over 10 independent replicates.
-For comparison, we also report the CPU time required by a
-modified version of the code used by \citet{martin2017} for a
-subset of these simulated ARGs.
+Supplementary Fig~\ref{fig:time}B shows that RAM usage is very
+modest for all these simulations.
 A comparison with existing methods that require reading
@@
 the entire dataset fits in RAM.
-Supplementary Fig~\ref{fig:time}B shows that RAM usage is very
-modest for all these simulations.
+To emphasise this point, we also generated
+quantitative traits for the simulations based on
+the French-Canadian pedigree~\citet{anderson2023}, discussed in
+the introduction. It took 60 seconds and XXX GB of RAM to
+simulate a trait with 1000 causal sites using \texttt{tstrait}.
+Using one byte per genotype, the full variant matrix of this
+dataset would require 140TB of RAM.
+% If we have space, mention the real-data sims
 
-% TODO this is to be rewritten with just a single example from FC
-% dataset and optionally also mentioning real data from 1000G (space
-% permitting).
-Figure \ref{fig:time} (B) shows simulation time of \texttt{tstrait} with more
-realistic data. For example, \texttt{tstrait} is used to simulate traits from
-2.7 million individuals in the French Canadian dataset, which is the largest
-known simulated genetic dataset \citep{anderson2023}. It took 60 seconds to
-simulate a trait with 1000 causal sites, making \texttt{tstrait} to be
-practical to simulate quantitative traits from this largest ever simulated
-dataset. The inferred ARG from the 1000 Genomes project \citep{kelleher2019} is
-also used to measure the computational performance.
+% Figure \ref{fig:time} (B) shows simulation time of \texttt{tstrait} with more
+% realistic data. For example, \texttt{tstrait} is used to simulate traits from
+% 2.7 million individuals in the French Canadian dataset, which is the largest
+% known simulated genetic dataset \citep{anderson2023}. It took 60 seconds to
+% simulate a trait with 1000 causal sites, making \texttt{tstrait} to be
+% practical to simulate quantitative traits from this largest ever simulated
+% dataset. The inferred ARG from the 1000 Genomes project \citep{kelleher2019} is
+% also used to measure the computational performance.
 

### 2023-11-22 paper.tex

@@
 Simulation is a critical component of GWAS method development, and
-generally consists of two steps: (1) generate a synthetic genetic
-variation dataset; and (2) simulate quantitative traits
-based on this dataset.
-(The two steps may sometimes be combined
-in a single simulation, or we may use real data instead of simulations in
-step (2), but the basic principles are still the same.)
+generally consists of two steps.
+Firstly, we must simulate some genetic variation. This is
+usually done either by model-based population genetic
+simulation using methods such as
+\texttt{msprime}~\citep{baumdicker2022efficient}
+or \texttt{SLiM}~\citep{haller2023},
+or by statistical resampling from existing datasets
+using methods like \texttt{HAPGEN2}~\citep{su2011hapgen2}
+or \texttt{HAPNEST}~\citep{wharrie2023hapnest}.
+Secondly, given this genetic variation dataset, we can
+simulate quantitative traits.
 The combined genetic variation and
@@
 \texttt{sim\_phenotype} function.
+The user can either specify a number of causal sites to be chosen randomly
+along the genome (one, by default), or can directly provide the causal
+sites as input. Combined with the detailed information about
+mutations recorded in a \texttt{tskit} ARG, this provides the ability
+to model population and haplotype specific effects.
 The return value \texttt{result} is an object encapsulating
@@
 \section{Conclusion}
-We believe that in the coming years, simulation in tree sequence encoding will
-attract more users in need of conducting genetic simulations with complex
-population structure. Multiple studies highlight the advantages of using the
-ARG data structure in GWAS \citep{link2023tree,nowbandegani2023extremely,zhang2023}.
-ARG data
-structures are becoming increasingly useful in genetic studies, as it is possible
-to accurately infer biobank-scale genealogies from sequencing data
-\citep{zhang2023}, and simulate large sample whole-genome sequences with
-spatiotemporal metadata \citep{anderson2023}. These advances in ARG presents a
-real opportunity for people to further understand the effects of complex
-population structure on quantitative traits. The \texttt{tstrait} package is an
-essential piece of infrastructure that lets people explore these possibilities.
+[WIP]
+There is a pressing need for large-scale truth datasets to support
+GWAS method development. One approach, exemplified by
+\citet{wharrie2023hapnest}, is to generate synthethic genetic variation
+based on publicly available data and to make these available. While
+there are certainly advantages to this method
+publicly
+
+For example, the \texttt{plink} formatted data file of chromosome 9
+for 1 million individuals provided for download is 78GB.
+
+% We believe that in the coming years, simulation in tree sequence encoding will
+% attract more users in need of conducting genetic simulations with complex
+% population structure. Multiple studies highlight the advantages of using the
+% ARG data structure in GWAS \citep{link2023tree,nowbandegani2023extremely,zhang2023}.
+% ARG data
+% structures are becoming increasingly useful in genetic studies, as it is possible
+% to accurately infer biobank-scale genealogies from sequencing data
+% \citep{zhang2023}, and simulate large sample whole-genome sequences with
+% spatiotemporal metadata \citep{anderson2023}. These advances in ARG presents a
+% real opportunity for people to further understand the effects of complex
+% population structure on quantitative traits. The \texttt{tstrait} package is an
+% essential piece of infrastructure that lets people explore these possibilities.
 

### 2023-11-23 paper.tex
Tweak abstract. Can't say "existing methods can't be applied" anymore

@@
 application areas, with ARG-based methods already providing substantial
-improvements over the state-of-the-art. The ability to efficiently
-simulate quantitative traits is critical to developing
-and evaluating new GWAS methods, but existing simulators do not scale
-to modern datasets and cannot be directly applied to ARGs.
-We present the \texttt{tstrait} package,
+improvements over the state-of-the-art.
+Sophisticated methods exist to simulate ARGs based on a variety
+of population genetic models, but there is currently no method
+to simulate quantitative traits directly from these ARGs.
+To apply existing quantitative trait simulators users must export
+genotype data, losing important information about ancestral processes
+contained in the ARG and producing large files when applied
+to Biobank-scale datasets.
+We present \texttt{tstrait},
 an open-source Python library to simulate
 quantitative traits on ARGs, and show how this user-friendly software
-can be easily applied to vast, population-scale datasets.
+can quickly simulate phenotypes for population-scale datasets
+on a laptop computer.
 \\\textbf{Availability and Implementation:} \texttt{tstrait} is available

### 2023-11-23 paper.tex
Progress

@@
 Many loci that are associated with various human diseases and
-traits have been identified~\citep{locke2015,ishigaki2022,
-mahajan2022,yengo2022,mathieson2023},
+traits have been
+identified~\citep[e.g.][]{yengo2022,mathieson2023},
+% \citep{locke2015,ishigaki2022,
+% mahajan2022,yengo2022,mathieson2023},
 and GWAS results are actively being incorporated into clinical
@@
 datasets~\citep{tanjo2021practical},
-but the sampling and resulting vast volumes of data
-are presenting significant challenges to GWAS
+but such scale is presenting significant challenges to GWAS
 methology~\citep{uffelmann2021}.
@@
 Simulation is a critical component of GWAS method development, and
-generally consists of two steps.
-Firstly, we must simulate some genetic variation. This is
+generally consists of first simulating some genetic
+variation, and then simulating quantitative traits based on it.
+The combined genetic variation and
+simulated quantitative traits then represent truth-data
+which association testing methods can be evaluated against.
+Simulating genetic variation is
 usually done either by model-based population genetic
@@
 or \texttt{HAPNEST}~\citep{wharrie2023hapnest}.
-Secondly, given this genetic variation dataset, we can
-simulate quantitative traits.
-The combined genetic variation and
-simulated quantitative traits then represent the ``truth'' against
-which association testing methods can be evaluated.
-The first step, of generating realistic, large-scale genetic variation
-datasets, is a well-studied problem with a mature and robust supporting
-software ecosystem. ARG-based methods~\citep{kelleher2016efficient,
-kelleher2018efficient,haller2018tree,adrion2020,baumdicker2022efficient,anderson2023}
-have been instrumental in making population genetic simulations
-of the scale and realism
-required to model modern GWAS cohorts~\cite[e.g.][]{martin2017,zaidi2020}.
-These simulation methods are all based on the ``succinct tree
-sequence'' ARG encoding~\citep{wong2023general},
-using the open-source \texttt{tskit} library~\citep{ralph2020}.
+Both approaches have advantages and disadvantages.
+Roughly speaking, model-based
+simulation methods provide better control of
+population processes such as demography but can be overly simplistic,
+whereas resampling methods better capture some of the nuance of
+real data, while lacking a direct connection to population genetics models.
+Model-based population genetic simulations have made
+great strides in recent years, with major advances in both
+scalability~\citep{kelleher2016efficient,kelleher2018efficient,haller2018tree}
+and realism~\citep{adrion2020,anderson2023}, and
+have been successfully used to model large-scale GWAS
+cohorts~\cite[e.g.][]{martin2017,zaidi2020}.
 
-An important property of the succinct tree sequence ARG encoding
-used by these population genetic simulators is that it allows us to
-store enormous genetic variation datasets compactly. Instead of
-storing the genetic sequence variation in the conventional
+An important part of these simulation methods is that they
+output ARGs using
+the ``succinct tree sequence'' encoding~\citep{wong2023general}
+and the open-source \texttt{tskit} library~\citep{ralph2020}.
+A key benefit of this approach is that it allows us to
+store enormous genetic variation datasets compactly.
+Instead of storing the genetic sequence variation in the conventional
 matrix-based form~\citep[e.g.\ VCF,][]{danecek2011},

### 2023-11-27 paper.tex

@@
 genotype data, losing important information about ancestral processes
-contained in the ARG and producing large files when applied
+and producing large files when applied
 to Biobank-scale datasets.
@@
 quantitative traits on ARGs, and show how this user-friendly software
-can quickly simulate phenotypes for population-scale datasets
+can quickly simulate phenotypes for Biobank-scale datasets
 on a laptop computer.
@@
 Simulation is a critical component of GWAS method development, and
-generally consists of first simulating some genetic
-variation, and then simulating quantitative traits based on it.
-The combined genetic variation and
-simulated quantitative traits then represent truth-data
-which association testing methods can be evaluated against.
-Simulating genetic variation is
-usually done either by model-based population genetic
-simulation using methods such as
+generally consists of two steps:
+first simulating genetic variation (genotypes)
+and then simulating quantitative traits (phenotypes) based on the
+synthetic genotypes.
+The combined genotypes and phenotypes represent ground-truth data
+which GWAS methods can be evaluated against.
+Genetic variation is
+usually simulated either by model-based population genetic
+methods such as
 \texttt{msprime}~\citep{baumdicker2022efficient}
@@
 or \texttt{HAPNEST}~\citep{wharrie2023hapnest}.
-Both approaches have advantages and disadvantages.
-Roughly speaking, model-based
-simulation methods provide better control of
-population processes such as demography but can be overly simplistic,
-whereas resampling methods better capture some of the nuance of
-real data, while lacking a direct connection to population genetics models.
+Both approaches have advantages and disadvantages and excel
+in different situations.
+Roughly speaking, model-based simulation methods
+provide better control of population processes such as demography,
+whereas resampling methods are better at capturing
+difficult to model nuances of real data.
 Model-based population genetic simulations have made
@@
 and realism~\citep{adrion2020,anderson2023}, and
-have been successfully used to model large-scale GWAS
+have been successfully used to simulate large-scale GWAS
 cohorts~\cite[e.g.][]{martin2017,zaidi2020}.
 
-An important part of these simulation methods is that they
-output ARGs using
-the ``succinct tree sequence'' encoding~\citep{wong2023general}
-and the open-source \texttt{tskit} library~\citep{ralph2020}.
-A key benefit of this approach is that it allows us to
-store enormous genetic variation datasets compactly.
-Instead of storing the genetic sequence variation in the conventional
-matrix-based form~\citep[e.g.\ VCF,][]{danecek2011},
-we can store the underlying genealogical history
-and mutations that ultimately \emph{result} in the observed
-variation~\citep{kelleher2019}.
+An important property of these modern simulation methods is that
+they output ARGs rather than genotypes \citep[e.g.\ as VCF,][]{danecek2011}.
+Using the succinct tree sequence ARG encoding~\citep{wong2023general}
+and the open-source \texttt{tskit} library~\citep{ralph2020},
+these ARG methods can simulate genetic variation for millions of
+samples and store the output very compactly.
 For example, a highly realistic simulation
@@
 French-Canadian individuals~\citep{anderson2023}
-requires around 550GB of storage space in
-gzip-compressed VCF ($\sim$280TB uncompressed).
-The original simulation file representing
-precisely the same genetic variation data, plus the
-complete simulated ARG, requires only 1.36GB of space
-in compressed \texttt{tszip} format. Thus,
-by encoding in terms of the underlying ARG,
-in this case we achieve more than 400-fold compression,
-and this highly unwieldy dataset
-becomes straightforward to work with (we discuss this
-landmark dataset again in later sections).
-Another substantial benefit of storing simulated datasets
-in this ARG-based format is that many computations can
+requires around 550GB of storage space in gzip-compressed VCF.
+At 1.36GB the original simulated ARG
+(compressed using the \texttt{tszip} utility) is around 400X smaller.
+Furthermore, many computations can
 be expressed efficiently in terms of the underlying
-genealogies~\citep{kelleher2016efficient,ralph2020}, without
+ARG~\citep{kelleher2016efficient,ralph2020}, without
 needing to decode the actual variation data.
+Finally, outputting a simulated ARG provides access to the full
+history, not just the genetic variation among the samples.
 
@@
 easy way to simulate quantitative traits based on such an ARG.
-Existing methods to simulate quantitative traits from a given set of
-genetic sequences in standard formats~\citep[e.g.][]{meyer2018,fernandes2020}
-are overwhelmed by the scale of the data. For example,
-just parsing the 280TB of VCF for the French-Canadian
-dataset mentioned above would take many days of CPU time,
-making quantitative trait simulation by this approach unfeasible.
+Many existing methods to simulate quantitative traits from a given set of
+genetic sequences assume that the genotypes fit in
+memory~\citep[e.g.][]{meyer2018,fernandes2020}, which makes them
+impractical at Biobank scale (e.g., the French-Canadian dataset discussed
+above would require 140TB of RAM assuming 1 byte per genotype).
+Methods that read parts of the genotype matrix from file as
+required~\citep[e.g.][]{wharrie2023hapnest} can be used on
+reasonable hardware, but working with such large files is slow
+and cumbersome.
+More fundamentally, exporting genotypes discards much of rich
+detail about ancestral history contained in an ARG, and
+it is exactly this information that we wish to take
+advantage of when using inferred ARGs in GWAS applications.
 In their analysis of the portability of polygenic risk
 scores across populations, \citet{martin2017}
-demonstrated the ability to simulate phenotypes at scale using ARGs.
-However, their approach is tightly coupled to the details
-of the study
-and not designed to be reused.
+demonstrated the utility of simulating phenotypes directly
+from an ARG. Their approach, however, is tightly coupled to the details
+of the study and not designed to be reused.
 Simulation code can be subtle and difficult to

### 2023-11-27 paper.tex
Pass

@@
 Ancestral nodes are not associated with individuals here,
-but in general the ARG may be embedded in a multigenerational
+but in general an ARG may be embedded in a multigenerational
 pedigree, where some internal nodes would be associated
@@
 \texttt{Tstrait} is a Python library, building on the \texttt{tskit}
-ARG library~\citep{ralph2020,wong2023general} and the rich
+ARG toolkit~\citep{ralph2020,wong2023general} and the rich
 Python data-science ecosystem~\citep{numpy}.
@@
 sites as input. Combined with the detailed information about
-mutations recorded in a \texttt{tskit} ARG, this provides the ability
-to model population and haplotype specific effects.
+mutations recorded in a \texttt{tskit} ARG, explicitly specifying causal
+sites allows us to model many different types of trait, for example
+those associated with mutations arising in a particular population
+or time interval.
 The return value \texttt{result} is an object encapsulating
@@
 The simulation results can then be efficiently and conveniently
-processed by a range of familiar Python data science tools.
+processed using familiar Python data science tools.
 
@@
 \texttt{tstrait} provides modular building blocks for power-users
-and potential integration with other tools that generate
+and to facilitate integration with other tools that generate
 traits on an ARG. The \texttt{sim\_trait} function
 simulates effect sizes for an input ARG, and returns a data frame
-describing the causal alleles and effect sizes.
+describing the causal sites, alleles and effect sizes.
 This dataframe can then be passed to the \texttt{genetic\_values}
@@
 causal mutations during the progress of a forwards-time simulation
-(to model the effect of selection, for example) could output these
+could output these
 values to a CSV or similar file. The modular architecture  and simple
@@
 modest for all these simulations.
-A comparison with existing methods that require reading
-in genetic sequences in formats such as
-VCF~\citep[e.g.][]{meyer2018,fernandes2020} is not feasible,
-as the VCFs involved are very large, and these tools assume that
-the entire dataset fits in RAM.
-To emphasise this point, we also generated
-quantitative traits for the simulations based on
-the French-Canadian pedigree~\citet{anderson2023}, discussed in
-the introduction. It took 60 seconds and XXX GB of RAM to
-simulate a trait with 1000 causal sites using \texttt{tstrait}.
-Using one byte per genotype, the full variant matrix of this
-dataset would require 140TB of RAM.
-% If we have space, mention the real-data sims
-
-% Figure \ref{fig:time} (B) shows simulation time of \texttt{tstrait} with more
-% realistic data. For example, \texttt{tstrait} is used to simulate traits from
-% 2.7 million individuals in the French Canadian dataset, which is the largest
-% known simulated genetic dataset \citep{anderson2023}. It took 60 seconds to
-% simulate a trait with 1000 causal sites, making \texttt{tstrait} to be
-% practical to simulate quantitative traits from this largest ever simulated
-% dataset. The inferred ARG from the 1000 Genomes project \citep{kelleher2019} is
-% also used to measure the computational performance.
+To demonstrate its scalability we  applied \texttt{tstrait} to the
+French-Canadian simulations~\cite{anderson2023} of chromosome 9
+discussed in the introduction.
+% TODO redo this for 100 causal sites @daiki.
+It took 6 seconds and XXX GB of RAM to
+simulate a trait with 100 causal sites using \texttt{tstrait}.
+% Is this worth the space? We should cite 1KG project also really.
+Finally, to demonstrate that \texttt{tstrait} can also be applied
+to ARGs inferred from real data, we simulated a trait with 100 causal
+sites for the ARG estimated from chromosome 9 of the 1000 Genomes
+project data~\citep{kelleher2019} which has YYY samples
+and ZZZ variant sites. This took 0.X seconds and Y GB of RAM.
 

### 2023-11-28 paper.tex
spell check

@@
 but such scale is presenting significant challenges to GWAS
-methology~\citep{uffelmann2021}.
+methodology~\citep{uffelmann2021}.
 
@@
 quantitative-trait locus mapping~\citep{link2023tree};
-and can provide a sparse and efficient model of linkage disequilibirium
+and can provide a sparse and efficient model of linkage disequilibrium
 in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

### 2023-11-28 paper.tex
Pass through discussion

@@
 % We need to check whether this actually works in the final version.
-\begin{lstlisting}[language=Python]
+\begin{lstlisting}[language=Python,aboveskip=1em,belowskip=1em]
 import tstrait as tst
@@
 Supplementary Fig~\ref{fig:time}A shows how trait simulation time scales
-with number of individuals, for sequences of length 50, 100 and 200Mb.
-For each sample size and sequence length,
-we first simulate an ARG under a simple coalescent model with
-human-like parameters using~\texttt{stdpopsim}~\citep{adrion2020}.
-We then run a quantitative trait simulation
-with 1000 causal sites on this ARG using \texttt{tstrait},
+with number of individuals on human-like coalescent simulations
+generated using~\texttt{stdpopsim}~\citep{adrion2020}.
+For each simulated ARG, we simulate a trait with 1000 causal sites
 and report the mean CPU time over 10 independent replicates.
 Supplementary Fig~\ref{fig:time}B shows that RAM usage is very
-modest for all these simulations.
-To demonstrate its scalability we  applied \texttt{tstrait} to the
-French-Canadian simulations~\cite{anderson2023} of chromosome 9
-discussed in the introduction.
+modest. To emphasise scalability, we also
+applied \texttt{tstrait} to the simulations of 1.4 million
+French-Canadians discussed in the introduction.
 % TODO redo this for 100 causal sites @daiki.
 It took 6 seconds and XXX GB of RAM to
-simulate a trait with 100 causal sites using \texttt{tstrait}.
+simulate a trait with 100 causal sites.
 % Is this worth the space? We should cite 1KG project also really.
@@
 to ARGs inferred from real data, we simulated a trait with 100 causal
-sites for the ARG estimated from chromosome 9 of the 1000 Genomes
-project data~\citep{kelleher2019} which has YYY samples
+sites for an ARG estimated from 1000 Genomes project
+data~\citep{kelleher2019} which has YYY samples
 and ZZZ variant sites. This took 0.X seconds and Y GB of RAM.
@@
 \section{Conclusion}
-[WIP]
-There is a pressing need for large-scale truth datasets to support
-GWAS method development. One approach, exemplified by
-\citet{wharrie2023hapnest}, is to generate synthethic genetic variation
-based on publicly available data and to make these available. While
-there are certainly advantages to this method
-publicly
-
-For example, the \texttt{plink} formatted data file of chromosome 9
-for 1 million individuals provided for download is 78GB.
-
-% We believe that in the coming years, simulation in tree sequence encoding will
-% attract more users in need of conducting genetic simulations with complex
-% population structure. Multiple studies highlight the advantages of using the
-% ARG data structure in GWAS \citep{link2023tree,nowbandegani2023extremely,zhang2023}.
-% ARG data
-% structures are becoming increasingly useful in genetic studies, as it is possible
-% to accurately infer biobank-scale genealogies from sequencing data
-% \citep{zhang2023}, and simulate large sample whole-genome sequences with
-% spatiotemporal metadata \citep{anderson2023}. These advances in ARG presents a
-% real opportunity for people to further understand the effects of complex
-% population structure on quantitative traits. The \texttt{tstrait} package is an
-% essential piece of infrastructure that lets people explore these possibilities.
+There is substantial interest in using inferred ARGs to improve
+association testing
+methods~\citep{zhang2023,link2023tree,nowbandegani2023extremely},
+and there is a pressing need for a well-tested, efficient
+and user-friendly means of simulating phenotypes on ARGs.
+Highly realistic simulations conditioned on large
+pedigrees~\citep{anderson2023} provide an exciting opportunity
+to test the effects of intricate population structure on
+GWAS, and we hope that \texttt{tstrait} will facilitate these
+investigations. \texttt{Tstrait}'s modular architecture
+and flexible specification of causal sites
+should provide the opportunity to explore new avenues of research,
+and an extensible platform for future development.
 

### 2023-11-28 paper.tex
Use journal abbreviations
Tweak abstract

@@
 genetics. Recent breakthroughs have made it possible to simulate and
-infer ARGs at the population scale, and there is now intense interest
-in applying ARG-based methods across a broad range of applications.
-Genome-wide association studies (GWAS) are one of the most important
-application areas, with ARG-based methods already providing substantial
-improvements over the state-of-the-art.
-Sophisticated methods exist to simulate ARGs based on a variety
-of population genetic models, but there is currently no method
+infer ARGs at biobank scale, and there is now intense interest
+in using ARG-based methods across a broad range of applications,
+particularly in genome-wide association studies (GWAS).
+Sophisticated methods exist to simulate ARGs using
+population genetics models, but there is currently no software
 to simulate quantitative traits directly from these ARGs.
@@
 genotype data, losing important information about ancestral processes
-and producing large files when applied
-to Biobank-scale datasets.
+and producing prohibitively large files when applied
+to the biobank-scale datasets currently of interest in GWAS.
 We present \texttt{tstrait},
@@
 quantitative traits on ARGs, and show how this user-friendly software
-can quickly simulate phenotypes for Biobank-scale datasets
+can quickly simulate phenotypes for biobank-scale datasets
 on a laptop computer.
@@
 templates is available on \url{https://tskit.dev/tstrait/docs/}, and the
-development version is maintained on Github
+development version is maintained on GitHub
 (\url{https://github.com/tskit-dev/tstrait}).\\ \textbf{Contact:}

### 2023-11-28 paper.tex
Edit pass. Shuffle some intro content

@@
 practice~\citep{visscher2017}.
-This success has led to the collection of many population-scale Biobank
-datasets~\citep{tanjo2021practical},
-but such scale is presenting significant challenges to GWAS
-methodology~\citep{uffelmann2021}.
-
-Ancestral recombination graphs (ARGs) encode the interwoven paths
-of genetic inheritance arising from
-% Explanation of citations: people might get offended if we don't cite
-% Griffiths when first mentioning the ARG, but Hudson has just as
-% much to do with the actual definition that we use in Wong 2023.
-recombination~\citep{hudson1983properties,griffiths1997ancestral,wong2023general},
-and recent
-breakthroughs in inference have opened many new possibilities for their
-practical application~\citep{lewanski2023era}.
-Building on these recent developments, substantial progress has been made
-on statistical genetics applications.
-For example,  ARG-based methods
-can detect more ultra rare variants than conventional association testing
-methods~\citep{zhang2023};
-have better power to detect causal loci in
-quantitative-trait locus mapping~\citep{link2023tree};
-and can provide a sparse and efficient model of linkage disequilibrium
-in GWAS and downstream applications~\citep{nowbandegani2023extremely}.
+The great success of GWAS has prompted the collection of many biobank
+datasets consisting of hundreds of thousands of
+participants~\citep{tanjo2021practical},
+but this scale presents significant challenges to
+current GWAS methodology~\citep{uffelmann2021}.
 
@@
 
-An important property of these modern simulation methods is that
-they output ARGs rather than genotypes \citep[e.g.\ as VCF,][]{danecek2011}.
-Using the succinct tree sequence ARG encoding~\citep{wong2023general}
-and the open-source \texttt{tskit} library~\citep{ralph2020},
-these ARG methods can simulate genetic variation for millions of
-samples and store the output very compactly.
+An important property of these population genetic simulation methods is that
+they output ancestral recombination graphs (ARGs)
+rather than sample genotypes.
+ARGs encode the interwoven paths
+of genetic inheritance arising from
+% Explanation of citations: people might get offended if we don't cite
+% Griffiths when first mentioning the ARG, but Hudson has just as
+% much to do with the actual definition that we use in Wong 2023.
+recombination~\citep{hudson1983properties,griffiths1997ancestral,wong2023general},
+and contain rich detail about ancestral processes.
+Recent breakthroughs in inferrence methods have made it possible
+to estimate ARGs at biobank scale~\citep{kelleher2019,zhang2023},
+and there is now intense interest in their practical
+application~\citep{lewanski2023era}.
+Statistical genetics has been a particular focus,
+and ARG-based methods have been shown
+to detect more ultra rare variants than conventional association testing
+methods~\citep{zhang2023};
+have better power to detect causal loci in
+quantitative-trait locus mapping~\citep{link2023tree};
+and to provide a sparse and efficient model of linkage disequilibrium
+in GWAS and downstream applications~\citep{nowbandegani2023extremely}.
+
+ARG-based methods can simulate genetic variation for millions of
+samples and store the output very compactly
+using the ``succinct tree sequence'' encoding~\citep{wong2023general}.
+and \texttt{tskit} library~\citep{ralph2020}.
 For example, a highly realistic simulation
@@
 French-Canadian individuals~\citep{anderson2023}
-requires around 550GB of storage space in gzip-compressed VCF.
+requires around 550GB of storage space in gzip-compressed
+VCF~\citep{danecek2011}.
 At 1.36GB the original simulated ARG
@@
 Although there are sophisticated methods available for simulating
-ARGs at the population scale, there is currently no
+ARGs at biobank scale, there is currently no
 easy way to simulate quantitative traits based on such an ARG.
@@
 memory~\citep[e.g.][]{meyer2018,fernandes2020}, which makes them
-impractical at Biobank scale (e.g., the French-Canadian dataset discussed
+impractical at biobank scale (the French-Canadian dataset discussed
 above would require 140TB of RAM assuming 1 byte per genotype).
@@
 
-In this paper we present \texttt{tstrait}, a new Python library that
-efficiently simulates quantitative traits on an arbitrary ARG
-in \texttt{tskit} format.
+In this paper we present \texttt{tstrait}, a Python library that
+efficiently simulates quantitative traits on an arbitrary ARG.
+% in \texttt{tskit} format.
 \texttt{Tstrait} can
 quickly simulate quantitative traits for population-scale datasets,
-with a very low memory overhead. This approach also provides a
-richer model than existing methods,
-allowing us to, for example, simulate traits for
-ancestral individuals as well as contemporary samples.
+with a very low memory overhead, and taking into account
+the rich historical detail contained within an ARG.
 The \texttt{tstrait} library also integrates well with the wider
